---
title: "Variable Scoping"
author: "Jasmine Hughes"
---

```{r setup, include=FALSE}
source("setup/_common.R")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Consider this notebook more an "FYI", the content won't be tested directly,
but understanding how scoping works may help you troubleshoot your code!

## Your Environment and Functions

Functions in R return an object. In general, R functions are and should be 
designed such that the only effect of the function is through the return value.

**Side effects** are when a function affects the state of the world in addition 
to its return value.  Can you think of any side effects that you saw an R 
function produce from earlier in the workshop?  What about:

- `library()`
- `setwd()`

Functions in R are (roughly) *pass-by-value* and not *pass-by-reference*. 
This means that if you modify an argument inside the function it will not change 
the original value outside the function. This protects you from a major 
potential source of side effects. (There are exceptions to this rule.)

In actuality, functions in R are *call-by-value*. What this means for our 
purposes is that you can pass an input argument in without a copy being made of 
it. This saves time and memory. At the time that you modify the input within the 
function (if ever), then a copy is made and  the modified input is different 
than the original value outside the function.

## Variable scope and global variables

In Lecture 1e, we saw that, when we call a function or other object, R looks for
what this object refers to in this order:

1. The global environment
2. Sequentially up the search path

Functions also create their own environment, which lasts until the function
finishes executing. When you refer to an object from within a function, the
order becomes:

1. The function environment
2. The global environment
3. Sequentially up the search path

Here are a couple examples to illustrate:

```{r}
# y is defined in the function:
add_to_y <- function(x) {
  y <- 8
  x + y
}

y <- 9
add_to_y(7)
```

In the example above, y is defined in the function (as 8) and so the value of 8
is used, even though the global value of y is 9.

```{r}
# y is defined outside the function (this is considered bad practice)
add_to_y <- function(x) {
  x + y
}

y <- 9
add_to_y(7) 
```

In this example, the function environment does NOT contain a definition of y and
so R looks for this object in the global environment, where it finds `y <- 9`.

There's one more (advanced) case that is useful to know about.

Functions are defined with a reference to the environment in which they are
created.

```{r}
# Part 1: defining a function within a function scope
add_numbers <- function(x) {
  y <- 10
  add_to_y <- function(z) {
    z + y
  }
  # call the function we just defined:
  return(add_to_y(z = x))
}

y <- 9
add_numbers(7)
```

What's going on?

When we call `add_to_y`, R looks in the environment in which that function was
defined, which is within the environment created by `add_numbers`, where 
`y <- 10` -- even though the global environment sets `y <- 9`.

Here's one more advanced example:

```{r}
# Part 2: defining a function within a global scope
y <- 10
add_to_y <- function(z) {
  z + y
}

# Define `y` within `add_numbers`
add_numbers <- function(x) {
  y <- 1000
  return(add_to_y(z = x))
}
add_numbers(7)
```

Here, `add_to_y` is defined in the *global environment*. When we call 
`add_to_y`, R looks for y in the following order:

1. The function environment of `add_to_y` --> no `y` is defined, so continue
2. The environment in which `add_to_y` is defined, the global environment.

If `y` did not exist in the global environment, R will next check our search
path (which contains all the packages we have laoded.)

We can remove `y` from the global environment, and R will *still not* use the
value `y <- 1000`.

## Breakout

In the following examples, try to predict what value `f(1)` will return 
*without* running the code.

1. What is `f(1)`?

```{r brk-1}
g <- function(x) {
  x + 1
}

f <- function(x) {
  g(x) + 1
}

f(1)
```

2. What will the following return?
(a) f(1)
(b) f(1, 3)

```{r brk-2}
g <- function(x, y = 1) {
  x + y
}

f <- function(x, y = 2) {
  g(x, y) 
}

f(1)
```

3. What is `f(1)`?

```{r brk-3}
g <- function(x, y = 1) {
  x + y
}
y <- 1000
f <- function(x) {
  g(x) 
}

f(1)
```
