---
title: "Loops!"
author: ["Jasmine Hughes", "Michael McCarthy"]
---

```{r setup, include=FALSE}
source("setup/_common.R")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Loops

Sometimes you need to repeat an operation iteratively. For this, we use loops.

There are three types:

* for: do something a set number of times: `for (x in 1:10) { }`
* while: do something until some condition becomes false: `while (x < 10) { }`
* repeat: keep repeating something until told to stop/break `repeat { }`

The last two have the potential to go on infinitely if not written correctly!
(Don't worry, you can always force R to stop...)

We will limit ourselves to `for` for now, but the other options can be very
useful in the right applications.

### The Syntax

Four parts:

1. Start with the word `for`, then provide brackets
2. In the brackets, define an increment variable (`i` is common)
3. In the brackets, define a set of possible options for the increment variable
4. Provide code for the body of the for-loop

```{r example_loop}
for (i in 1:10){
  print(paste("the value of i is", i))
}
```

The iterations do not have to explicitly be over sequential numbers.

```{r, for-nonsequential}
vec <- seq(0, 12, 2)
for (i in vec) {
 print(paste("the value of i is", i))
}
```

Nor do you need to iterate over numbers.

```{r nonnumeric}
for (lab in c("bilirubin", "creatinine", "WBC")){
  print(lab)
}
```

If you want to store the result of each loop somewhere, you need to initialize
the object *before* the loop starts. (Otherwise you'll keep overwriting your
variable.)

Let's write a for-loop to convert heights in centimeters to inches:

```{r for_versus_vectorized}
heights_cm <- c(165, 180, 150, 172)
height_in <- c()                     # initialize this object, which will grow

for (h in heights_cm) {
  val <- h / 2.54 
  height_in <- c(height_in, val)     # append the new value to our variable
}

height_in
```

How would we do this without for loops?

```{r}
height_in <- heights_cm / 2.54
height_in
```

Which approach is faster? Nearly always, the vectorized approach is (without the
for-loop).

Let's do a little experiment.

```{r}
heights_cm <- sample(140:190, 50000, replace = TRUE) # a big vector!

vectorized_cm2in <- function(heights_cm) {
  heights_cm / 2.54
}

forloop_cm2in <- function(heights_cm) {
  heights_in <- c()
  for (h in heights_cm) {
    heights_in <- c(heights_in, h/2.54)
  }
  heights_in
}

system.time(vectorized_cm2in(heights_cm))
system.time(forloop_cm2in(heights_cm))
```

There are things we can do to make our loops *faster*.

In the loops above, we created an empty vector and appended to it each time. 
This is quite slow in R. It's faster to create a vector of the size you need,
and insert values into it.

```{r}
# for loop, initializing the output vector to the final length:
faster_forloop_cm2in <- function(heights_cm) {
  heights_in <- rep(NA, length(heights_cm))
  for (h in 1:length(heights_cm)) {
    heights_in[h] <- heights_cm[h]/2.54
  }
  heights_in
}

system.time(faster_forloop_cm2in(heights_cm)) # orders of magnitude faster!
```

In many languages, looping (for loops, while loops, etc.) is one of the main 
constructs used to carry out computation. Loops are not emphasized as much in R, 
both because they can be slow and because other syntax is often cleaner.

However, loops can be very powerful and sometimes they are the right approach:

* Need to use a function that doesn't support vectorized operations (and you 
  cannot re-write it)
* The loop option is easier to read than the alternative, and
* Code performance is a low priority or dataset is small

## Beyond loops: `lapply` and `purrr::map`

Applying the same function to each element of a vector or list is such a common
task that R gives you tools built exactly for it. Instead of initializing an
output object and writing a `for` loop, you *map* a function over the input and
get the results back directly---usually with less code and fewer chances to
make a mistake.

### Base R: the `apply` family

`lapply()` ("list apply") applies a function to each element and returns a
**list**:

```{r}
heights_cm <- c(165, 180, 150, 172)

lapply(heights_cm, function(h) h / 2.54)
```

`sapply()` ("simplify apply") does the same but *simplifies* the result to a
vector when it can, which is often what you want:

```{r}
sapply(heights_cm, function(h) h / 2.54)
```

(`vapply()`, `mapply()`, and `apply()` are close relatives for related jobs.)

### The tidyverse: `purrr::map`

The `purrr` package (part of the tidyverse) provides a more consistent set of
mapping functions. `map()` always returns a list, while the typed `map_*()`
variants return a specific kind of vector. Here we write `purrr::` to make it
clear where the functions come from:

```{r}
purrr::map(heights_cm, function(h) h / 2.54)     # a list
purrr::map_dbl(heights_cm, function(h) h / 2.54) # a numeric vector
```

`map_chr()`, `map_lgl()`, and `map_int()` likewise return character, logical, and
integer vectors---asking for the type up front is a nice guard against surprises.

For this toy example the vectorized `heights_cm / 2.54` is still the simplest and
fastest option. `lapply()` and `map()` earn their keep when the operation *isn't*
vectorized---fitting a model to each patient's data, reading in a folder full
of files, or running a simulation for each set of parameters---exactly the
situations where you might otherwise reach for a loop.

## Breakout

1. Use a for loop to count the number of vowels in the alphabet. (see the 
variable `letters`) 

```{r brk-1}
# Hint: 
"a" %in% c("a", "e", "i", "o", "u")  # TRUE
"b" %in% c("a", "e", "i", "o", "u")  # FALSE
letters

```

2. Loop through all the integers between 1 and 100 (inclusive). If the value is
divisible by 5, print "fizz". If the value is divisible by 7, print "buzz". If
the value is divisible by both, print "fizzbuzz".

```{r brk-2}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0

```

3. Update your solution so that now instead of printing a message, you update a
counter. How many numbers between 1 and 100 (inclusive) are divisible by at 
least one of 5 or 7?

```{r brk-3}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0

```
