---
title: "Writing functions"
author: ["Jasmine Hughes", "Michael McCarthy"]
---

```{r setup, include=FALSE}
source("setup/_common.R")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Overview

So far, most of the tasks we've carried out in R have been basic operations, or 
have leveraged the `tidyverse` and other packages for plotting or working with 
data frames. However, sometimes you might need to perform somewhat more 
complicated tasks. In this lecture, we will look at some of the basic building 
blocks that you can put together to accomplish these tasks.

These are some of the key pieces missing from SAS, Stata, SPSS and other tabular 
data analysis software.

## Functions

Functions are one of the most important constructs in R (and many other 
languages). They allow you to modularize your code - encapsulating a set of 
repeatable operations as an individual function call.

You should rely heavily on functions rather than having long sets of expressions 
in R scripts.

Functions have many important advantages:

* They reduce bugs by avoiding having multiple instances of the same 
  functionality.
  
* They reduce time involved in coding by eliminating redundancy.

* They make for cleaner and more easily-readable code.

A basic goal in writing functions is *modularity*.

In general, a function should

* be fairly short,

* be focused and specific in what it does, and

* be designed so that it can be used in combination with other functions to 
  carry out more complicated operations.
  
Aspirational goals:

* No more than 20 lines long
* No more than 5 arguments

(A lot of core R functions fail these rules!)

## Writing functions

### The Basic Syntax

Create a function:

```{r}
# variable_name <- function(arg1, arg2) {
  # code goes here!
#}

add_two <- function(x){
  x + 2
}

add_two(4)
```

A function takes 0 or more arguments, and will return the last unassigned line 
of code, as above, or you can more explicitly `return()` a value.

```{r}
add_two <- function(x){
  y <- x + 2
  return(y)
}

add_two(4)

add_two <- function(x){
  return(x + 2)
}

add_two(4)
```

See how R changes the color of `function` and `return`? These are "special"
words.

### A more complicated example

Let's say that body mass index (BMI) is something you often have to calculate. 
You could remember the formula and type it out each time: bmi = weight/height**2

But it might be less error prone to write a function! 

```{r}
calculate_bmi <- function(height_in_cm, weight_in_kg){
  # Fill in the function code here!
}
```

Lets try it out!

```{r}
library(dplyr)

starwars |>
  select(name, height, mass) |>
  mutate(bmi = calculate_bmi(height_in_cm = height, weight_in_kg = mass))
```

## Function arguments

R can match arguments by name (when provided) or by position (the fall-back). 

```{r}
# supply arguments in correct order, use argument names --> good
starwars |>
  select(name, height, weight = mass) |>
  mutate(bmi = calculate_bmi(height_in_cm = height, weight_in_kg = weight))

# supply arguments in correct order, don't use argument names --> same answer
starwars |>
  select(name, height, weight = mass) |>
  mutate(bmi = calculate_bmi(height, weight))

# switch the argument order, but use argument names  --> same answer
starwars |>
  select(name, height, weight = mass) |>
  mutate(bmi = calculate_bmi(weight_in_kg = weight, height_in_cm = height))

# switch the argument order, but DON'T use argument names --> wrong answer!!
starwars |>
  select(name, height, weight = mass) |>
  mutate(bmi = calculate_bmi(weight, height))
```

It also allows one to specify default values so that the user doesn't have to 
explicitly provide all the arguments.

Let's update our `calculate_bmi()` function to round to some number of decimal
places, with a default of 2, via a new `n_decimals` argument.

```{r}
calculate_bmi <- function(height_in_cm, weight_in_kg) { # Update the arguments!
  # Update your function code here!
}
```

Write the function so these calls work:

```{r}
calculate_bmi(height_in_cm = 175, weight_in_kg = 70)
calculate_bmi(height_in_cm = 175, weight_in_kg = 70, n_decimals = 5)
calculate_bmi(height_in_cm = 175, weight_in_kg = 70, n_decimals = 0)
```

## What is the "..." argument for?

Using `...` as one of the arguments to a function allows a function to pass along user-provided arguments without specifying explicitly what the user might
provide. 

We can update our calculate_bmi function to make use of this behaviour:

```{r}
calculate_bmi <- function(height_in_cm, weight_in_kg, ...){
  height_in_m <- height_in_cm/100
  bmi <- weight_in_kg / height_in_m ** 2
  round(bmi, ...)
}

calculate_bmi(height_in_cm = 175, weight_in_kg = 70)
calculate_bmi(height_in_cm = 175, weight_in_kg = 70, digits = 5)
calculate_bmi(height_in_cm = 175, weight_in_kg = 70, digits = 0)
```

*Important*: when you pass arguments into a function via `...` the argument names must match!

## Breakout

1) Write an R function that will take an input vector and set any values less 
than zero to zero.

```{r brk-1}
# Write your solution here:


# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
out <- thresh0(vec)
identical(expect_out, out)
```

2) Update your answer so that your function takes a second argument for the 
threshold, so that instead of 0, the user inputs the value of the lower
threshold.

```{r brk-2}
# Write your solution here:

# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out1 <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
expect_out2 <- c(2, 2, 6, 2, 2)   # the same as vec with all values < 2 set to 2
out1 <- thresh(vec, 0)
out2 <- thresh(vec, 2)
identical(expect_out1, out1)
identical(expect_out2, out2)
```

3. Update your solution so that, by default, the lower threshold is 0, but the
user can override it by passing a lower threshold.

```{r brk-3}
# Write your solution here:

# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out1 <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
expect_out2 <- c(2, 2, 6, 2, 2)   # the same as vec with all values < 2 set to 2
out1 <- thresh(vec)               # only one argument provided
out2 <- thresh(vec, 2)            # both arguments are provided
identical(expect_out1, out1)
identical(expect_out2, out2)
```
