---
title: "Flow Control: if and else"
author: "Jasmine Hughes"
---

```{r setup, include=FALSE}
source("setup/_common.R")
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(dplyr)
library(datasets)
```

We have used logical expressions before:

```{r}
# logical expressions for subsetting
vec <- c(1, 5, -7, 13)
x <- vec > 0
y <- vec < 0 | vec >= 13

vec[x]
vec[y]

# logical expressions for subsetting in dplyr
filter(airquality, is.na(Solar.R))
```

Often you want to use logical expressions to control what code is evaluated.

For this, we will use `if` and `else`!

## if-else syntax

Here's a simple example to illustrate the syntax. 

```{r, if}
val <- rnorm(1)
if (val < 0) {
  print("val is negative")
} else {
  print("val is positive")
}
```

Note how RStudio colors these words differently? They are special key words.

We can chain together `if` statements using `else if `.

```{r, if-chain}
val <- rnorm(1, mean = 0, sd = 1)
val
if (val < -1) { 
  print("val is more than one standard deviation below the mean.")
} else if (abs(val) <= 1) {
  print("val is within one standard deviation of the mean.")
} else {
  print("val is more than one standard deviation above the mean.")
}
```

In general, the `{` brackets are only needed if you have multiple R expressions,
but R will complain when an `else` starts a line of code, so generally using the
`{` is good practice. That said, this _works just fine_:

```{r, if-oneline}
if (val < 0) print("val is negative") else print("val is positive")

# better:
if (val < 0) {
  print("val is negative") 
} else {
  print("val is positive")
}
```

## The condition in an if statement

There are two rules for the `if () ` statement:

1. The value evaluated must be TRUE or FALSE

   - NA, NULL, and NaN will all produce (different) errors
   
2. The value evaluated must be of length 1

   - If more than one value is provided, depending on your user settings, either
     the first value will be evaluated or you will encounter an error
   - If less than one value (`c()` is of length 0), you will encounter an error

Zero evaluates to `FALSE`, all other numbers evaluate to `TRUE`. In general
strings are not converted to booleans.

```{r}
# These will both produce errors
vec <- c(4, 8)
vec < 6
if (vec < 6) {
  print("vec is less than 6")
} else {
  print("vec is not less than 6")
}

vec <- NA
if (vec < 6) {
  print("vec is less than 6")
} else {
  print("vec is not less than 6")
}
```

If you anticipate missing data, a more robust alternative is to use `isTRUE()`:

```{r}
isTRUE(NA)
isTRUE(NULL)
isTRUE(1)
isTRUE(FALSE)
isTRUE(TRUE)
isTRUE(as.logical(1))

vec <- NA
if (isTRUE(vec < 6)) {
  print("vec is less than 6")
} else {
  print("vec is not less than 6")
}
```

If you want to be able to handle vectors, `all` or `any` are helpful:

```{r}
# `any` checks for at least one TRUE
vec <- c(4, 8)
if (any(vec < 6)) {
  print("at least one value is less than 6")
} else {
  print("no values are less than 6")
}

# `all` requires every value to be TRUE
vec <- c(4, 8)
if (all(vec < 6)) {
  print("all values are less than 6")
} else {
  print("at least one value is not less than 6")
}
```

## Vectorized if/else

Sometimes it is helpful to use this same `if...else...` logic in a vectorized
way, i.e., to apply different operations on a vector according to some condition

For this, we can use `ifelse`.

For example, we have a number of patient lab results, and we want to indicate
which lab results were normal versus elevated:

```{r alternative_one_line_if}
bilirubin <- data.frame(
  pt = c("M.H.", "P.J.", "K.G."),
  lab = c(1, 1.8, 0.7)
)

bilirubin
```

Elevated bilirubin is defined as a value above 1.2:

```{r}
mutate(
  bilirubin,
  bilirubin_test = ifelse(lab > 1.2, "elevated", "normal")
)
```

## Breakout

1. We previously saw a function for calculating BMI, which looked something like
this:

```{r brk-1}
calculate_bmi <- function(height, weight){ # assume height in m, weight in kg
  weight / height ** 2
}

calculate_bmi(1.70, 70)
```

Modify the function so that the user supplies, as a string, the unit for height,
accepting either c("cm", "m", or "in"). Then, if necessary, convert the height
to the appropriate unit before calculating BMI.

Use the following conversion factors:
1m = 100 cm = 39.37 inches

```{r brk-2}
# write your solution here

# if it is correct, the following should return TRUE
identical(calculate_bmi(1.70, 70, "m"), calculate_bmi(170, 70, "cm"))
identical(
  round(calculate_bmi(1.70, 70, "m"), 3), 
  round(calculate_bmi(1.7*39.37, 70, "in"), 3)
)

```
