---
title: "The Basics"
author: ["Jasmine Hughes", "Michael McCarthy"]
---

```{r setup, include=FALSE}
source("setup/_common.R")
knitr::opts_chunk$set(echo = TRUE)
```

## R as a calculator

Write some example expressions for each of the math operations below:

```{r computing,echo=TRUE,results='markup'}
# add numbers using +
2 + 5

# multiply numbers using *
20 * 3

# powers: two ways! `**` or `^`
2**2
2^2

# R uses math orders of operation: control with parentheses: ()
(10 / 2) + 10
```

Other common mathematical operations use *functions*:

```{r computing2,echo=TRUE,results='markup'}
# calculate a square root using `sqrt`
sqrt(16)

# calculate the remainder using `%%`
17 %% 5

# calculate a logarithm using `log`
log(10)
```

What base does R assume you are using by default? How do we change the base?

```{r}
?log
log(10, base = 10)
```

R also recognizes scientific notation:

```{r}
5000000000 * 1000
5e9 * 1e3
```

## Assigning values to R objects

So far we have been mostly coding "interactively"---running a single command and 
looking at the answer---or we have been saving objects (like graphs saved as 
pictures) to our "working directory".

But what if we want to use the same object for later calculations?

A key action in R is to store values in the form of R objects, and to examine 
the value of R objects. (Add notes to help remind yourself what's going on!)

```{r assigning_values}
# (variable name) <- (value, or code to create a value/object)
x <- 5

# (variable name) = (value, or code to create a value/object)
# equal sign *can* be used for assignment but is not conventional
y = 10
```

Notice how the values we saved weren't printed out?

```{r}
# In both the console and in R chunks,
# running a variable name will print its value out
x

# We can also use the function `print`
print(x)
```

Variable names have a few rules:

- can't start with a number
- case sensitive
- can contain some special characters: `.`, `_`

```{r}
x7 <- 7
X7 <- 8
x.y <- 9
x_9 <- 10
```

You can over-write a variable once you've created it:

```{r}
x7 <- "seven"
x7
```

You can see these values change in the "Environment" tab in Rstudio!

## Vectors in R

The most basic form of an R object is a vector. In fact, individual (scalar) 
values are vectors of length one. The variables we created earlier are all 
vectors!

We can concatenate values into a vector with `c()`.

```{r}
# Use `c()` with numbers
c(1, 2, 3)

# Use `c()` with character strings
c("a", "b", "c")
```

R also has tools for defining repeatable patterns. Keep it DRY: "Don't Repeat Yourself".

```{r sequences}
# Use `:` to increment by integers
1:10

# Or use `seq` for more control
?seq
seq(0, 20, by = 5)
```

### Numeric vectors

Also known as "double precision floating point".

```{r types_of_vectors1}
seq(10, 25, 5)
```

### Integer vectors

Less common than numeric vectors, but it's good to know they exist since
sometimes you'll see errors related to the difference between numeric and 
integer vectors.

```{r types_of_vectors2}
# Create an integer by putting `L` at the end. 
x1 <- 1
x2 <- 1L
typeof(x1)
typeof(x2)
```

Integers are not numbers! But they will often be coerced to numbers.

```{r}
identical(1L, 1)
1L == 1
typeof(1L + 1)
```

R Trivia: 'L' stands for 'long integer' (historical)

### Character vectors

Most common way of storing non-numeric data. 

```{r types_of_vectors3}
# create a character vector of your names using `c`
c("Alice", "Bob", "Carol")
```

Character strings must be wrapped in matching double quotes `"` or single quotes `'` (not a mix), or else R will try to interpret it as a variable. Double quotes are the conventional choice.

```{r}
# Note: the backslash is an "escape character"
chars <- c(
  'hi', 'hallo', "mother's", 'father\'s',  "She said, 'hi'", "He said, \"hi\""
)
chars
```

Often we want to combine multiple character strings together:

```{r}
# use the function `paste0`
paste0("Hello, ", "world")
```

### Logical vectors (or boolean vectors)

Logical values are either `TRUE` or `FALSE`. Type them without quotes, all-caps.

```{r types_of_vectors4}
TRUE

# create a logical vector
c(TRUE, FALSE)
c(T, F) # You *can* do this, but *don't*!
```

Logical values are often created from numerical comparisons:

```{r}
vals <- c(1, 8, 10)

# Less than: <
vals < 8

# Less than or equal to: <=
vals <= 8

# Greater than: >
vals > 8

# Greater than or equal to: >=
vals >= 8

# Equal: ==
vals == 8

# Not equal: !=
vals != 8
```

Make sure you use double equal signs for equality! Remember, single equal signs are for assigning objects:

```{r}
v1 <- 7
v2 <- 9
v1 == v2

v1 = v2
v1 <- v2 # same as the line above
identical(v1, v2)
```

What if you combine data types?

```{r types_of_vectors5}
# create a combination vector:
combo <- c(TRUE, 7, "frogs")

# How do I know what datatype something is? function: `typeof`
typeof(combo)
```

## Breakout 1

In general, your answers to any questions should involve writing code to 
manipulate objects. Don't open excel or another calculator to answer them.

1) Create a variable called `my_numbers` that contains the integers 1 through 10.
How many different ways can you think of doing this? Which do you find the
easiest to read? Which do you find the easiest to write?

```{r brk-1}
# your code goes here!

```

2) "Bobby has 3 apples. Susie has 7 apples. Susie has more apples than Bobby."
How would you communicate these facts in R code?

```{r brk-2}

```

## Working with indices and subsets

What if we have a vector of values but we only want to use certain elements?
Use square brackets to refer to particular positions in a vector:

```{r indexing}
vals <- seq(2, 12, by = 2)
vals
vals[1]
x <- vals[1]
x
```

Can you guess what these will do before running the code??

```{r indexing2}
vals
vals[3:5]
vals[c(1, 3)]
vals[-c(1, 3)]
```

We can also use logical values!

```{r}
which_indices <- c(rep(TRUE, 3), rep(FALSE, 2), TRUE)
which_indices
vals[which_indices]
vals == 6
vals[vals == 6]
vals[(vals %% 4) == 0]
```

## Comparing Values

We already saw numeric comparisons:

```{r comparisons}
vec1 <- c(1, 2, 3, 4, 5)
vec2 <- c(7, -4, 3, 4, 5.1)
vec2 < vec1
vec1 < vec2
vec1 == vec2
vec1 != vec2
```

We can also do other types of comparisons:

```{r}
## using 'or': |
vec1 >= 3 | vec1 < -7

## using 'and': &
vec1 >= 3 & vec1 %% 2 == 0 
```

What are the results of each of these?

```{r}
TRUE | FALSE
FALSE | FALSE
TRUE & FALSE
TRUE & TRUE
```

What answer do you expect here?

```{r}
"apple" < "banana"
x <- "apple"
"apple" == x
```

## Breakout 2

In general, your answers to any questions should involve writing code to 
manipulate objects. Don't open excel or another calculator to answer them.

3) Create a vector of the squares of all numbers from 1-10. Save it to a
variable called `x`

```{r brk-3}
# your code goes here!

```

4) Use a logical expression to identify which elements of `x` are less than 20.
Save this logical vector as a variable called `y`.

```{r brk-4}

```

5) Use the vector `y` to subset x to return (a) all the elements less than 20 
and (b) all the elements greater than or equal to 20.

```{r brk-5}
# (a)

# (b)

```

6) How would you interpret the results of these expressions?

```{r brk-6}
x
sum(x)
mean(x)

y
sum(y)
mean(y) 
```
