---
title: 'Lecture 2, solutions'
author: "Jasmine Hughes"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# Breakout 2a

1. Use `dplyr` and the `starwars` data set. Create a new data frame containing
only droid characters (hint: `species` column), and all columns except
`vehicles` and `starships`.

```{r}
dplyr::starwars |>
  filter(species == "Droid") |>
  select(-vehicles, -starships)

```

2. How many Star Wars characters are from Tatooine and have a height of at least 
100 cm?

```{r}

# Option 1
starwars |>
  filter(
    homeworld == "Tatooine",
    height > 100
  ) |>
  nrow()

# Option 2
starwars |>
  filter(homeworld == "Tatooine") |>
  filter(height > 100) |>
  nrow()

# Option 3
starwars |>
  filter(homeworld == "Tatooine" & height > 100) |>
  nrow()

```

3. Use `dplyr` to sort the `starwars` data frame alphabetically by name.

```{r}
starwars |>
  arrange(name)

arrange(starwars, name)
```

4. Use `dplyr` and `grepl()` to filter for characters with the name "Skywalker".

```{r}
starwars |>
  filter(grepl("Skywalker", name))

```


# Breakout 2b

1. Use `dplyr` to create a data frame containing the maximum temperature for 
each month in the `airquality` data set.

```{r}
airquality |>
  group_by(Month) |>
  summarize(
    max_temp = max(Temp)
  )

```

2. Use `dplyr` to add a column to the `airquality` dataset that contains the
average temperature in that month.

```{r}
airquality |>
  group_by(Month) |>
  mutate(
    avg_temp = mean(Temp)
  )
```

3. Use `dplyr` to add a column called `delta_from_avg` that contains the
difference between that day's `Temp` and the mean `Temp` of that month. Arrange
the dataframe by the column you just created, in descending order (so that the
relatively hottest days are listed first).

```{r}
airquality |>
  group_by(Month) |>
  mutate(
    avg_temp = mean(Temp),
    delta_from_avg = Temp - avg_temp
  ) |>
  arrange(desc(delta_from_avg))
  
```

4. Use `dplyr` to create a data frame counting the number of days with a Temperature above 80F in each month. (You might find the function `n` helpful.)

```{r}
airquality |>
  filter(Temp > 80) |>
  group_by(Month) |>
  summarize(n_days = n())
```

# Breakout 2c
1. Create a factor variable with the levels of the factor in the order of the 
colors of the rainbow ("red", "orange", "yellow", "green", "blue", "indigo", 
"violet"). Create a second factor variable with the levels of the factor in 
alphabetical order.

```{r}
rainbows <- c("red", "orange", "yellow", "green", "blue", "indigo", "violet")

```

2. Below are two different datetimes. How many hours apart are they? How many
days apart are they? Use an argument supplied to `difftime` to convert units,
rather than dividing by 24.
```{r}
dt1 <- "2024-03-14 11:00:00"
dt2 <- "2024-03-23 19:17:00"

difftime(as.POSIXct(dt2), as.POSIXct(dt1), unit = "hours")

difftime(as.POSIXct(dt2), as.POSIXct(dt1), unit = "days")

```

3. Demonstrate, using the function `identical`, that setting a column in
`test_data_frame` to NULL is the same as using `dplyr::select` to remove a
column from a data frame.

```{r}
test_data_frame <- mtcars

# set a column to NULL
test_data_frame$mpg <- NULL

# remove a column with select
dplyr_solution <- mtcars |>
  select(-mpg)

identical(test_data_frame, dplyr_solution)
```

4. Coerce the vector `my_numbers` to a numeric vector. Count how many NA's were
introduced. (Hint: R interprets TRUE as having a value of 1 and FALSE as having
a value of 0)

```{r}
my_numbers <- c('23', 'z', '37', '69', 'h', '18', '18', '19', '91', '4', '97', '83', '19', '51', '93', '20', '80', '37', '34', '93', '36', '55', 't', '48', '27', '54', '3', '89', '40', '73', '44', '52', '78', 's', '89', '84', '18', '59', '89', '14', '77', '52', '62', '92', 'o', '45', '1', '81', '3', '7', '24', '57', '50', '56', '95', '85', '100', '92', '98', '60', '18', '55', '2', '11', '8', '45', 'a', '6', '77', '9', '49', '66', '78', '18', '21', '36', 'i', '58', '45', '27', '13', '47', '54', '92', '86', '9', '94', '58', '62', 'k', '87', '34', '41', '58', '48', '12', '76', '31', '24', '20', '7', '40', '24', '98', '86', '51', '53', '29', '87', '20', '15', '97', '59', '20', '32', '61', '34', '92', '31', '36', '8', '49', '92', '9', '55', '10', '26', '100', '70', '77', '92', '18', '20', '10', '26', '4', '1', '5', '11', '99', '98', '2', '45', '24', '7', '98', '79', '64', '79', '15', '43', '20', '9', '58', '85', '66', '41', '16', '16', '94', '62', '21', '53', '43', '63', '4', '33', '97', '47', '73', '72', '38', '97', '71', '55', '66', '50', '91', '96', '21', '64', '70', '23', '8', '38', '27', '18', '9', '92', '43', '35', '23', '22', '17', '43', '42', '8', '42', '69', '7', '95', '54', '39', '24', '100', '61', '65', '16', '33', '48', '72', '94', '19', '68', '42', '52', '51', '83', '55', '21', '64', '29', '76', '56', '55', '55', '52', '42', '80', '94', '78', '46', '58', '33', '21', '30', '33', '60', '50', '22', '55', '27', '22', '65', '54', '31', '5', '10', '1', '21', '62', '90', '36', '38', '80', '100', '62', '75', '56', '29', '74', '44', '99', '45', '98', '16', '57', '11', '34', '35', '47', '95', '4', '36', '26', '74', '10', '44', '52', '85', '31', '4', '11', '62', '83', '27', '53', '7', '45', '43', '13', '9', '11', '7', '34', '87', '65', '19', '28', '17', '51', '1', '18', '19', '95', '52', '82', '53', '4', '5', '87', '27', '11', '69', '5', '6', '39', '52', '26', '1', '13', '98', '80', '66', '68', '11', '100', '57', '58', '64', '85', '49', '90', '12', '5', '47', '1', '99', '47', '65', '98', '88', '2', '64', '98', '57', '10', '64', '16', '79', '91', '91', '67', '58', '61', '49', '49', '56', '37', '61', '43', '20', '34', '76', '99', '67', '50', '32', '63', '82', '16', '94', '62', '27', '83', '77', '43', '41', '12', '57', '49', '55', '3', '81', '61', '39', '48', '23', '72', '95', '64', '49', '42', '81', '66', '30', '35', '80', '95', '45', '38', '34', '40', '4', '34', '34', '91', '80', '42', '26', '69', '100', '46', '6', '32', '99', '43', '66', '47', '20', '1', '67', '14', '77', '89', '19', '42', '3', '92', '34', '55', '98', '79', '74', '90', '6', '76', '84', '91', '90', '88', '53', '13', '75', '21', '96', '12', '38', '54', '7', '37', '57', '27', '89', '81', '31', '54', '21', '93', '60', '23', '96', '43', '48', '13', '30', '57', '9', '50', '29', '27', '60', '56', '53', '18', '71', '97', '44', '73', '93', '96', '34', '44', '19', '8', '62', '23', '68', '28', '13', '90', '40', '25', '45', '53', '86', '60', '100', '70', '17')

numeric_numbers <- as.numeric(my_numbers)
sum(is.na(numeric_numbers))

```

