---
title: 'Lecture 1, solutions'
author: "Jasmine Hughes"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```



# Breakout 1c
In general, your answers to any questions should involve writing code to 
manipulate objects. Don't open excel or another calculator to answer them.

1) Create a variable called `my_numbers` that contains the integers 1 through 10.
How many different ways can you think of doing this? Which do you find the
easiest to read? Which do you find the easiest to write?

```{r}
# this works well, and it's very clear, but takes a while to type
# (and you might make a typo!)
my_numbers <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)

# both of these are clear and concise, it's a matter of personal taste!
my_numbers <- 1:10
my_numbers <- seq(1, 10)

# overly creative, but the end result is the same.
# (you might find such solutions are harder to read)
my_numbers <- c(1, 2, 3:6, seq(7, 10))

# very, very clear about outputs (rather than using defaults)
my_numbers <- seq(from = 1, to = 10, by = 1)

```

2) "Bobby has 3 apples. Susie has 7 apples. Susie has more apples than Bobby."
How would you communicate these facts in R code?

```{r}
bobby_apples <- 3
susie_apples <- 7
susie_gt_bobby_apples <- susie_apples > bobby_apples
susie_gt_bobby_apples 
```

3) Create a vector of the squares of all numbers from 1-10. Save it to a
variable called `x`


```{r}
x <- (1:10) ** 2
```

4) Use a logical expression to identify which elements of `x` are less than 20.
Save this logical vector as a variable called `y`.

```{r}
y <- x < 20
```

5) Use the vector `y` to subset x to return (a) all the elements less than 20 
and (b) all the elements greater than or equal to 20.

```{r}
# (a)
x[y]

# (b)
x[!y]

```

6) How would you interpret the results of these expressions?
```{r}
x 
sum(x)  # sum of squares from 1-10
mean(x) # mean of squares from 1-10

# since x is numeric, these expressions behave like we might mathematically
# expect

y
sum(y)  # number of values less than 20
mean(y) # proportion of values less than 20

# since y is a logical, mathematical operations convert TRUE to 1 and FALSE to 0
# `sum` calculates 1 + 1 + 1 + 1 + 0...
# `mean` calculates the mean (sum(y)/length(y)) which is in this case also the
# proportion of TRUE values (since all values of y are either 1 or 0)

```

# Breakout 1d

In general, your answers to any questions should involve writing code to 
manipulate objects. For example, if I ask you to find the maximum temperature, 
do not scan through all the values and find it by eye. Use R to do the 
calculations and print results.

Some of these questions will require you to search for the right function.

**Hint**: use the function `apropos("key word")` to find something to use.

1) The variable x contains all the temperature measurements from the air quality
data frame. Create a variable called 'avg_temp' that contains the mean 
temperature. Then, create a vector of length 2 that contains the minimum and 
maximum temperature and assign it to a variable with a meaningful name.
```{r}
library(datasets)
x <- airquality$Temp
x

# write your solution below
avg_temp <- mean(x)
temp_extremes <- c(min(x), max(x))

```

2) Create a boolean (TRUE/FALSE) vector indicating whether the temperature in x 
is higher than the average temperature.

```{r}
x > avg_temp
```

3) The temperature recordings are in Fahrenheit. Convert the column to Celsius.
Save the column to the data frame as a new column called "TempC"
Hint: C = (F - 32) * 5/9

```{r}
airquality$TempC <- (airquality$Temp - 32) * 5/9
airquality
```

4) Use functions in R to round 'TempC' to two decimal places and to two 
significant digits.

```{r}
round(airquality$TempC, 2)
signif(airquality$TempC, 2)
```

# Breakout 1 e

1) Make sure you are able to install packages from CRAN. E.g., try to install 
*praise*. What does the code below do?

```{r }
install.packages("praise")
praise::praise()
# This function returns motivational messages to the user.

```

2) Figure out what your current working directory is.

```{r }
getwd()
```

3) Write the theophylline data frame to your downloads folder. Open it in excel 
or a similar application. Does it look like you expected?

```{r }
write.csv(
  datasets::Theoph,
  file.path("~", "Downloads")
)

```



