---
title: "Practice Problems"
output: html_document
date: "2024-02-27"
---

Unless otherwise specified, these questions ask you to use the provided data,
(gapminder.csv) which contains 6 columns and 1704 rows. 

The columns are as follows:

* country: Name of country
* year: Year (YYYY, stored as a number),
* pop: Population of country in given year
* continent: Continent of country
* lifeExp: Life expectancy (years) of country in given year
* gdpPercap: GDP per capita, in international dollars, fixed price.

# dplyr

1. Use `dplyr` to create a data frame containing data from before 1975 with 
every column except `continent`.

```{r}

```

2. How many countries had a life expectancy greater than 60 in the year 1982?

```{r}

```

3. Use `dplyr` and `grepl()` to filter for countries with the letter "z" in their name.

```{r}


```

4. Use `dplyr` to sort the gap data set in reverse alphabetical order.

```{r}


```

# ggplot

1) Plot a histogram of life expectancy. 

```{r}

```

2) Plot the life expectancy against gdpPercap using `geom_point`

```{r}

```

3) Clean up your scatterplot with a title and axis labels. Output it as a PDF 
and see if you'd be comfortable with including it in a report/paper.

```{r}


```

4) Make a boxplot of life expectancy (y) versus year (x).

```{r}


```

5) Use `facet_wrap` to create a trellis plot of life expectancy by gdpPercap
scatterplots, one subplot per year.  

```{r}

```

6) Plot life expectancy versus gdpPercap. Now plot so that different continents 
are in different colors. Use `scale_x_continuous()` to set the x-axis limits to 
be in the range from 100 to 50000.

```{r}

```

7) Figure out how to use the log-scale for gdpPercap, without manually 
calculating the log values.

```{r}

```

8) Create a "trellis" plot (using `facet_wrap`) where, for a given year, each 
panel uses a) hollow circles to plot lifeExp as a function of log(gdpPercap), 
and b) a red loess smoother without standard errors to plot the trend. Turn off 
the grey background. Figure out how to use partially-transparent points to 
reduce the effect of the overplotting of points.

```{r}

```

9) Building on your graph in Question 8, change at least two different 
properties in the `theme()` layer to customize the graph to your liking.

```{r}

```

# Programming

1) Write a for loop that will loop through a vector of values and produce a
vector equal to the input vector except with negative values set to zero. (ex:
`c(-2.3, 0, 6, 9) should become c(0, 0, 6, 9))

```{r}


```

2. Below is a list of instructions that control a robot. The robot can move
up or down only. The "+" instruction indicates the robot moves up, and the "-"
instruction indicates the robot moves downwards. The robot starts at a position of zero. What position does the robot end at?

```{r}
instr <- c(
  "-", "-", "+", "+", "-", "+", "+", "-", "-", "-", "-", "+",
  "+", "-", "-", "+", "+", "-", "+", "-", "-", "-", "+", "-", "-",
  "-", "-", "-", "-", "-", "-", "-", "-", "-", "-", "+", "-", "+",
  "-", "+", "-", "-", "+", "-", "+", "-", "-", "-", "+", "-", "-",
  "-", "-", "+", "-", "+", "-", "+", "+", "-", "+", "-", "-", "-",
  "-", "+", "-", "-", "-", "-", "-", "+", "+", "+", "-", "-", "+",
  "+", "+", "+", "+", "-", "+", "+", "+", "+", "-", "-", "-", "+",
  "-", "-", "-", "-", "+", "+", "+", "+", "-", "-", "-", "-", "-",
  "-", "+", "-", "-", "+", "-", "-", "+", "+", "+", "-", "+", "-",
  "-", "+", "-", "+", "-", "+", "+", "-", "+", "-", "-", "+", "+",
  "-", "+", "-", "-", "+", "-", "-", "+", "+", "+", "-", "-", "-",
  "+", "+", "+", "+", "-", "-", "+", "-", "-", "-", "+", "+", "-",
  "+", "-", "+", "-", "-", "-", "+", "-", "+", "-", "+", "-", "-",
  "-", "-", "+", "-", "-", "+", "-", "-", "+", "-", "+", "-", "-",
  "-", "-", "+", "+", "+", "+", "+", "-", "-", "-", "-", "+", "+",
  "-", "+", "+", "+", "+", "-", "-", "-", "+", "+", "+", "-", "+",
  "-", "+", "+", "+", "-", "-", "-", "-", "-", "-", "-", "-", "+",
  "+", "-", "-", "-", "-", "-", "-", "+", "+", "+", "+", "+", "+",
  "+", "-", "+", "-", "-", "-", "+", "+", "+", "+", "+", "+", "+",
  "-", "+", "-", "+", "+", "-", "+", "+", "+", "-", "+", "-", "-",
  "-", "-", "+", "+", "+", "+", "-", "+", "-", "+", "-", "-", "+",
  "-", "-", "+", "-", "-", "+", "-", "+", "+", "-", "+", "+", "+",
  "+", "-", "+", "+", "-", "+", "+", "+", "-", "+", "-", "-", "-",
  "+", "+"
)


```
