---
title: "Practice Problems"
output: html_document
date: "2024-02-27"
---

Unless otherwise specified, these questions ask you to use the provided data,
(gapminder.csv) which contains 6 columns and 1704 rows. 

The columns are as follows:

* country: Name of country
* year: Year (YYYY, stored as a number),
* pop: Population of country in given year
* continent: Continent of country
* lifeExp: Life expectancy (years) of country in given year
* gdpPercap: GDP per capita, in international dollars, fixed price.

# dplyr

1. Use `dplyr` to create a data frame containing data from before 1975 with 
every column except `continent`.

```{r}
gap %>%
  filter(year < 1975) %>%
  select(-continent)

```

2. How many countries had a life expectancy greater than 60 in the year 1982?

```{r}
gap %>%
  filter(year == 1982) %>%
  filter(lifeExp > 60) %>%
  nrow()


gap %>%
  filter(
    year == 1982,
    lifeExp > 60
  ) %>%
  nrow()
```

3. Use `dplyr` and `grepl()` to filter for countries with the letter "z" in their name.

```{r}
?grepl

gap %>%
  filter(grepl("Z|z", country)) %>%
  arrange(desc(country))

gap %>%
  filter(grepl("z", tolower(country)))  %>%
  arrange(desc(country))

gap %>%
  filter(grepl("z", country, ignore.case=TRUE))  %>%
  arrange(desc(country))

```

4. Use `dplyr` to sort the gap data set in reverse alphabetical order.

```{r}
gap %>%
  arrange(desc(country))


```

# ggplot

1) Plot a histogram of life expectancy. 

```{r}
hist(gap$lifeExp)

gap %>%
  ggplot() +
  aes(x = lifeExp) +
  geom_histogram(binwidth = 5)
```

2) Plot the life expectancy against gdpPercap using `geom_point`

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y =lifeExp) +
  geom_point()
```

3) Clean up your scatterplot with a title and axis labels. Output it as a PDF 
and see if you'd be comfortable with including it in a report/paper.

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y =lifeExp) +
  geom_point(size = 0.1, alpha = 0.5) +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  scale_x_log10() +
  theme_classic()

```

4) Make a boxplot of life expectancy (y) versus year (x).

```{r}
gap %>%
  ggplot() +
  aes(y = lifeExp, x = factor(year)) +
  geom_boxplot()

```

5) Use `facet_wrap` to create a trellis plot of life expectancy by gdpPercap
scatterplots, one subplot per year.  

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y =lifeExp) +
  geom_point(size = 0.1, alpha = 0.5) +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  scale_x_log10() +
  theme_classic() +
  facet_wrap(~year) +
  theme(
    strip.background = element_blank()
  )

```

6) Plot life expectancy versus gdpPercap. Now plot so that different continents 
are in different colors. Use `scale_x_continuous()` to set the x-axis limits to 
be in the range from 100 to 50000.

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y =lifeExp, color = continent) +
  geom_point(size = 0.3, alpha = 0.5) +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  scale_x_continuous(limits = c(100, 50000)) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )
```

7) Figure out how to use the log-scale for gdpPercap, without manually 
calculating the log values.

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y =lifeExp, color = continent) +
  geom_point(size = 0.3, alpha = 0.5) +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  scale_x_log10(limits = c(100, 50000)) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  )
```

8) Create a "trellis" plot (using `facet_wrap`) where, for a given year, each 
panel uses a) hollow circles to plot lifeExp as a function of log(gdpPercap), 
and b) a red loess smoother without standard errors to plot the trend. Turn off 
the grey background. Figure out how to use partially-transparent points to 
reduce the effect of the overplotting of points.

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y = lifeExp) +
  geom_smooth(method = "loess", color = "red") +
  geom_point(alpha = 0.5, size = 0.2) +           # alpha controls transparency
  facet_wrap(~year) +
  scale_x_log10() +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  theme_bw()

```

9) Building on your graph in Question 8, change at least two different 
properties in the `theme()` layer to customize the graph to your liking.

```{r}
gap %>%
  ggplot() +
  aes(x = gdpPercap, y = lifeExp) +
  geom_smooth(method = "loess", color = "red") +
  geom_point(alpha = 0.5, size = 0.2) +           # alpha controls transparency
  facet_wrap(~year) +
  scale_x_log10() +
  theme_bw() +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy (years)",
    title = "Relationship between GDP and life expectancy"
  ) +
  theme(
    strip.background = element_rect(fill = "white"),
    axis.ticks = element_line(color = "black")
  )
```

# Programming

1) Write a for loop that will loop through a vector of values and produce a
vector equal to the input vector except with negative values set to zero. (ex:
`c(-2.3, 0, 6, 9) should become c(0, 0, 6, 9))

```{r}
vec <- c(-2.3, 0, 6, 9)
out <- c()

for (v in vec) {
  if (v < 0) {
    v <- 0
  }
  out <- c(out, v)
}
identical(out, c(0, 0, 6, 9))

```

2. Below is a list of instructions that control a robot. The robot can move
up or down only. The "+" instruction indicates the robot moves up, and the "-"
instruction indicates the robot moves downwards. The robot starts at a position of zero. What position does the robot end at?

```{r}
instr <- c(
  "-", "-", "+", "+", "-", "+", "+", "-", "-", "-", "-", "+",
  "+", "-", "-", "+", "+", "-", "+", "-", "-", "-", "+", "-", "-",
  "-", "-", "-", "-", "-", "-", "-", "-", "-", "-", "+", "-", "+",
  "-", "+", "-", "-", "+", "-", "+", "-", "-", "-", "+", "-", "-",
  "-", "-", "+", "-", "+", "-", "+", "+", "-", "+", "-", "-", "-",
  "-", "+", "-", "-", "-", "-", "-", "+", "+", "+", "-", "-", "+",
  "+", "+", "+", "+", "-", "+", "+", "+", "+", "-", "-", "-", "+",
  "-", "-", "-", "-", "+", "+", "+", "+", "-", "-", "-", "-", "-",
  "-", "+", "-", "-", "+", "-", "-", "+", "+", "+", "-", "+", "-",
  "-", "+", "-", "+", "-", "+", "+", "-", "+", "-", "-", "+", "+",
  "-", "+", "-", "-", "+", "-", "-", "+", "+", "+", "-", "-", "-",
  "+", "+", "+", "+", "-", "-", "+", "-", "-", "-", "+", "+", "-",
  "+", "-", "+", "-", "-", "-", "+", "-", "+", "-", "+", "-", "-",
  "-", "-", "+", "-", "-", "+", "-", "-", "+", "-", "+", "-", "-",
  "-", "-", "+", "+", "+", "+", "+", "-", "-", "-", "-", "+", "+",
  "-", "+", "+", "+", "+", "-", "-", "-", "+", "+", "+", "-", "+",
  "-", "+", "+", "+", "-", "-", "-", "-", "-", "-", "-", "-", "+",
  "+", "-", "-", "-", "-", "-", "-", "+", "+", "+", "+", "+", "+",
  "+", "-", "+", "-", "-", "-", "+", "+", "+", "+", "+", "+", "+",
  "-", "+", "-", "+", "+", "-", "+", "+", "+", "-", "+", "-", "-",
  "-", "-", "+", "+", "+", "+", "-", "+", "-", "+", "-", "-", "+",
  "-", "-", "+", "-", "-", "+", "-", "+", "+", "-", "+", "+", "+",
  "+", "-", "+", "+", "-", "+", "+", "+", "-", "+", "-", "-", "-",
  "+", "+"
)

# very succinctly:
ifelse(instr == "+", 1, -1) |> sum()

# in a for-loop
pos <- 0
for (i in instr) {
  if (i == "+") {
    pos <- pos + 1
  } else {
    pos <- pos - 1
  }
}
pos

```
