---
title: 'Introduction to ggplot2'
author: "Jasmine Hughes"
date: "12/01/2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# Why ggplot2?

ggplot2's elegance and customizability has made it the most popular way to 
visualize data in R.

Let's load the packages we need and read in our data sets we made earlier:

```{r}
library(ggplot2)
library(dplyr)
medal_counts <- read.csv(file.path("..", "data", "medal_counts.csv"))
medal_counts_wide <- read.csv(file.path("..", "data", "medal_counts_wide.csv"))
usa_gold_medals <- read.csv(file.path("..", "data", "usa_gold_medals.csv"))

```

The general call for `ggplot2` graphics looks something like this:

```{r, eval=FALSE}
# This code will not work, it is pseudo code to demonstrate the concept
library(ggplot2)
ggplot(data = ) +
  geom_xxxx(aes(x = ,y = , [options]))) + 
  geom_xxxx(aes(x = ,y = , [options]))) + 
  ... + 
  ... + 
  ...
```

Note that `ggplot2` graphs in layers in a *continuing call* (hence the endless 
+...+...+...), which really makes the extra layer part of the call.

**common mistake!** Careful: `+` is not the same as `|>` (common typo!) We are 
adding layers together, not piping the input of one function into another.

You can see the layering effect by comparing the same graph with different 
colors for each layer:

```{r, fig.cap=" ", warning=FALSE}
p <- ggplot(data = medal_counts_wide) +
  aes(x = Year, y = Gold) +                   
  geom_point(color = "gold")
p
p + 
  geom_point(
    aes(x = Year, y = Silver), 
    color = "gray"
  ) + 
  ylab("Medals")
```

# Grammar of Graphics

`ggplot2` syntax is very different from base graphics and lattice. It's built on 
the **grammar of graphics**. The basic idea is that the visualization of all 
data requires four items:

1) A dataset and set of mappings from variables to **aesthetics** 
   (ex: year -> x, medal count ->y). In ggplot, use `aes()`

2) A **geometry** (`geom_xxx`), which describes how the aesthetics data is 
   transformed into the graphical representation in terms of **kind** 
   (or **shape**).

3) One **scale** for each aesthetic mapping used. Scales describe how the 
   aesthetics data is transformed into the graphical representation in terms of 
   **degree**. (What does a value of `5` or `10` when determining what each 
   pixel in a graph looks like?). 

   * `scale_color_manual` and family for mapping values to colors/linetypes/
     circle 
     types (displayed in a figure legend)
   * data transofmrations (log, etc), described using gridlines and axes for
     position scales

4) A coordinate system, affects all position variables (cartesian, 
   logarithmic/semi-log, polar)


*Aside*: "The Grammar of Graphics" is a book, if you're interested in data 
visualization theory!

```{r, warning = FALSE, message = FALSE, fig.cap = "", fig.height=2.5}
medal_counts_wide |>
  ggplot() +
  aes(x = Year, y = Gold) +
  geom_point() +
  ggtitle("Gold Medal Counts") +
  labs(y = "N gold medals")

```

```{r}

ggplot(usa_gold_medals, aes(x = Year, y = count)) + 
  geom_line() +
  ggtitle("USA Gold Medals")

```

```{r}

# Boxplots
ggplot(medal_counts_wide, aes(x = factor(Year), y = Gold)) +
  geom_boxplot() + 
  ggtitle("Gold Medal Counts")
```

```{r}
# Histogram
ggplot(medal_counts_wide, aes(x = Gold)) + 
  geom_histogram() +
  ggtitle("Gold Medal Counts")
```

```{r}
# Density plot
ggplot(medal_counts_wide, aes(x = Gold)) + 
  geom_density() +
  ggtitle("Gold Medal Counts")
```

```{r}
# Bar chart
ggplot(usa_gold_medals, aes(x = Year, y = count)) + 
  geom_bar(stat = "identity")
```

* Consistent API call
* Change out one `geom` for another to change the "type"

# The ordering of layers matters!

We saw that we can add multiple layers on top of each other, but how does
ggplot know which one to do first?

```{r layeroverride}

# Compare boxplot first, then jitter
iris |>
  ggplot() +
  aes(x = Species, y = Sepal.Length) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.3)  # jitter: point with position dodging

# versus jitter first, then boxplot
iris |>
  ggplot() +
  aes(x = Species, y = Sepal.Length) +
  geom_jitter(width = 0.3) +
  geom_boxplot(outlier.shape = NA)


```

# `ggplot2` and tidy data

`ggplot2` plays nice with `dplyr` and pipes. If you want to manipulate your 
data specifically for one plot but not save the new dataset, you can call 
your `dplyr` chain and pipe it directly into a `ggplot` call.

```{r, fig.cap = "", fig.show = "hold"}
# This combines the subsetting and plotting into one step
medal_counts |>
  filter(Medal == "Gold") |>
  filter(NOC == "USA") |>
  ggplot(aes(x = Year, y = count)) + 
  geom_line()

```

# Exporting your graph

### The best way: ggsave

ggsave ensures your graph is the same size every time you create it:

```{r, eval=FALSE}
?ggsave

ggsave(
  filename = "example.pdf", 
  plot = p, 
  width = 7,
  height = 7,
  bg = "white"
)
```

### A convenient way: plots panel

Great for "quick" graphs, easy sharing. Not great for automatically generating 
graphs, reproducibly generating a graph.

# Breakout Problems

These are the same problems from before, but now try solve them with ggplot2!

1. Using the airquality dataset, plot a histogram of temperature.

```{r}
datasets::airquality
```

2. Extend your previous answer to add a plot title, and label the axes.
(Hint! We saw the function `labs()` before.)
```{r}

```

3. Create a new directory called `figures`. Save your plot to the `figures` 
directory.

```{r}

```

4. Using the ToothGrowth dataset, plot a box plot of tooth length by dose and
supplement. 
(Hint! You might need to make a new column or variable for the x-axis)
```{r}
datasets::ToothGrowth

```
