---
title: 'Managing your R Environment'
author: "Jasmine Hughes"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

# The Global Environment

The global environment is your workspace. It includes all the objects you've
created, which are listed in the "Environment" tab.

Every time you call a function, a new environment is created, with its own set
of objects. This environment lasts until the function finishes running.

We'll see more about this in Lecture 4.


## What's in the workspace?

R has functions for learning about the collection of objects in your workspace.
Some of this is built in to RStudio.

```{r}
x <- rnorm(5)
y <- c(5L, 2L, 7L)
z <- list(a = 3, b = c('sam', 'yang'))
ls()  # search the user workspace (global environment)
rm(x)    # delete a variable
```


# Packages (R's killer app)

Let's check out the [packages on CRAN](https://cran.r-project.org/web/packages/) 

Essentially any well-established and many not-so-established statistical methods 
and other functionality is available in a package. 

If you want to sound like an R expert, make sure to call them *packages* and 
not *libraries*. A *library* is the location in the directory structure where 
the packages are installed/stored.

### Using packages

Two steps:

1. Install the package on your machine
2. Load the package

To install a package, in RStudio, just do `Packages->Install Packages`.

From the command line, you generally will just do
```{r eval = FALSE}
install.packages('dplyr') 
?install.packages
library(dplyr)
```
That should work without specifying the repository from which to download the 
package (though sometimes you will be given a menu of repositories from which to 
select) but sometimes you'll get errors indicating a package is not available 
for your version of R if you don't include an explicit repository. 

If you're on a network and are not the administrator of the machine, you may 
need to explicitly tell R to install it in a directory you are able to write in:
```{r eval = FALSE}
install.packages('dplyr', lib = file.path('~', 'R'))
```

Now would be a good point to install the various packages we need for the 
workshop, which can be done easily with the following command:

```{r eval=FALSE}
install.packages(
  c(
    'datasets',     # useful for example problems
    'tidyverse',    # all sorts of data science tools!
    'ggplot2',      # graphing!
    'haven',        # reading/writing specialized data types
    'scales',       # customized graph axes
    'gridExtra',    # combine multiple plots into one
    'reprex',       # reproducible examples
    'xlsx',         # working with excel
    'knitr',        # working with R Markdown notebooks
    'PKPDsim'       # simulate concentration-time curves given PK parameters
  )
)
```

Note that packages often are dependent on other packages so these dependencies 
may need to be installed first. This should happen automatically, but sometimes
you might get prompted that there was a problem installing a package because
some dependency was missing.

You can also install directly from a package zip/tarball rather than from CRAN 
by giving a filename instead of a package name.

### General information about a package

You can use syntax as follows to get a list of the objects in a package and a 
brief description: `library(help = packageName)`. 

```{r}
library(help = "datasets")
```

On CRAN there often *vignettes* that are an overview and describe usage of a 
package if you click on a specific package. The *reference manual* is just a 
single document with the help files for all of the objects/functions in a 
package, so may be helpful but often it's hard to get the big picture view from 
that.

```{r}
vignette("base", package = "dplyr")
```

More on `dplyr` in the next lecture!

# Packages: The Technical Details

Environment: a set of objects (funcions, data, ...) available for you to use.

Package: a set of objects (usually functions, but also other objects) available
for you to use.

What's the difference?

### The search path

If you use an object, R will search in this order:

1. The function environment (if called from within a function).
2. The global environment.
3. The search path.

Loading a package adds it to your search path.

The order in which you load packages changes the search path priority!

To see the packages that are loaded and the order in which packages are searched 
for functions/objects: `search()`.

```{r search_path_experiment}
# 1. Identify packages currently loaded
search()
original_search <- search()

# 2. Load a new package
library(lubridate) # a tidyverse package
lubridate::now()

# 3. How did our search path change?
search_after_library_load <- search()
setdiff(search_after_library_load, original_search)
```

You can "reset" your search path to your default by "Refreshing" your R Session.

```{r search_path_experiment_part2}
search_after_refresh <- search()
setdiff(search_after_library_load, search_after_refresh)
```

**Tip**: Start every analysis and homework problem with a fresh session and a 
clean environment! Out-of-date objects and package dependencies you forgot to
declare are common sources of bugs or reproducibility issues.

(*Reproducibility*: Other people should be able to run your code/analysis and
get the same results.)

How does R know where to look for packages when you call `library`?

```{r}
.libPaths() # functions and variables that start with "." 
            # are usually for "advanced" users.
```

### Package namespaces

A package namespace contains all the objects that package makes available for
you to use.

Namespaces are way to keep all the names for objects in a package together in a 
coherent way and allow R to look for objects in a principled way.

A few useful things to know:

```{r}
library(dplyr)
# You can reference functions within a package using two colons:
stats::t.test(rnorm(15, 6), rnorm(20, 7))

# This is useful when two functions have the same name:
?filter
?dplyr::filter
?stats::filter

```

### Creating your own R package

R is do-it-yourself - you can write your own package! At its most basic this is 
just some R scripts that are packaged together in a convenient format. And if 
giving it to someone else, it's best to have some documentation in the form of 
function help files. 

Why make a package?

* It's an easy way to share code with collaborators
* It's a good way to create self-contained code for code you commonly use yourself
* It's how you can share your code and methods with the outside world
* It helps make your work reproducible
* It forces you to be more formal about your coding, which will improve your code

See the *devtools* package and *usethis* package for some useful tools to 
help you create a package. And there are lots of tips/tutorials online, in 
particular [Hadley Wickham's R packages book](https://r-pkgs.had.co.nz/).

Here are a few packages I've made:
- https://github.com/insightRX/clinPK: clinical calculations
- https://github.com/insightRX/PKPDsim: simulate PK/PD curves
- https://github.com/insightRX/PKPDposterior: Full Bayes estimation for MIPD 

### When should I create my own package?

* If you use the same code more than once: make a function (Lecture 4)
* If you use the same function in more than one project or assignment, 
  make a package.
* You don't need to be an expert to make a package!

# Breakout 1

1) Make sure you are able to install packages from CRAN. E.g., try to install 
*praise*. What does the code below do?

```{r }
praise::praise()

```

# The working directory

To read and write from R, you need to have a firm grasp of where in the 
computer's filesystem you are reading and writing from. 

R Notebooks (like this one) always run from the file location. 

The Console working directory can be configured, either from the console, or
through the UI.

```{r, eval=FALSE}
getwd()  # what directory will R look in?
# Linux/Mac specific
setwd('~/Documents') # change the working directory
setwd('/Users/jasminehughes/Documents') # absolute path
getwd()


# Windows - use either \\ or / to indicate directories
# setwd('C:\\Users\\Your_username\\Desktop\\R_for_pharmacometricians')
# setwd('..\\R_for_pharmacometricians')

# use `file.path` to be platform-agnostic.
```

Many errors and much confusion result from you and R not being on the same page
in terms of where in the directory structure you are.

```{r}
list.files()
```


# Reading text files into R

The workhorse for reading into a data frame is `read.table()`, which allows any 
separator (CSV, tab-delimited, etc.). `read.csv()` is a special case of 
`read.table()` for CSV files.

Earlier we used `read.csv()` to load pharmacokinetic data for theophylline.

```{r}
theophylline <- read.csv(
  file.path("..", "example_project", "data", "theophylline.csv")
)

theophylline
```

It's good to first look at your data in plain text format outside of R and then 
to check it after you've read it into R.

Careful about handling [genomic information in excel](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1044-7)!

# Reading Excel sheets into R

Excel is widely used, and so data often comes in Excel spreadsheets. There are
a number of packages for reading Excel's proprietary formats. I like `readxl`.

```{r}
require(readxl)
aq <- readxl::read_excel(file.path("..", "data", "airquality.xlsx"))
aq
```

# Reading 'foreign' format data

Other statistical software tools often produce specialized file types. The
package `haven` works well for these:

* SAS: read_sas() reads .sas7bdat + .sas7bcat files
* SPSS: read_sav() reads .sav files; read_por() reads .por files.
* Stata: read_dta() reads .dta files. 

```{r}
library(haven)
?haven
```

R can also read in (and write out) Excel files, netCDF files, HDF5 files, etc., 
in many cases through add-on packages from CRAN. 

**Tip**: Best practices for data-sharing: avoid using proprietary formats. 
TXT and CSV files are much easier to read across platforms and programming 
languages.

# Writing data out from R

Here you have a number of options. 

1) You can write out R objects to an R Data file using `saveRDS()`.
2) You can use `write.csv()` and `write.table()` to write data frames to flat 
   text files with delimiters such as comma and tab.
3) You can use `write()` to write out matrices in a simple flat text format.
4) You can use `cat()` to write to a file, while controlling the formatting.
5) You can write out in the various file formats mentioned above


# Breakout 2

2) Figure out what your current working directory is.

```{r }

```

3) Write the theophylline data frame to your downloads folder. Open it in excel 
or a similar application. Does it look like you expected?

```{r }
datasets::Theoph

```


4) (Optional) If you have a .csv file or a Microsoft Excel file with data that 
you've used for an unrelated project or assignment, try open it now, just to see 
if you can!

```{r}

```

