---
title: "Lecture 3 Solutions"
output: html_document
---

# Breakout Problems - 3a

1. Using the airquality dataset, plot a histogram of temperature.

```{r}
datasets::airquality
hist(datasets::airquality$Temp)
```

2. Using the ToothGrowth dataset, plot a box plot of tooth length by dose and
supplement. (Hint! You can use the "formula" argument to combine multiple 
variables)
```{r}
as.formula("y ~ x1 + x2") # example
boxplot(len ~ supp + dose, data = datasets::ToothGrowth)
```

# Breakout Problems - 3b

These are the same problems from before, but now try solve them with ggplot2!

1. Using the airquality dataset, plot a histogram of temperature.

```{r}
datasets::airquality |>
  ggplot() +
  aes(x = Temp) +
  geom_histogram()
```

2. Extend your previous answer to add a plot title, and label the axes.
(Hint! We saw the function `labs()` before.)
```{r}
p <- datasets::airquality |>
  ggplot() +
  aes(x = Temp) +
  geom_histogram(binwidth = 5) +
  labs(
    x = "Temperature (F)",
    y = "Number of days",
    title = "Temperature between May and September"
  )
```

3. Save your plot to the `figures` directory.

```{r}
ggsave(
  file.path("..", "figures", "temperature_histogram_ggplot.pdf"),
  p
)
```

4. Using the ToothGrowth dataset, plot a box plot of tooth length by dose and
supplement. 
(Hint! You might need to make a new column or variable for the x-axis)
```{r}
datasets::ToothGrowth |>
  mutate(xlabel = paste0(supp, "_", dose)) |>
  ggplot() +
  aes(y = len, xlabel) +
  geom_boxplot()

```

# Breakout Questions - 3c

1) The datasets::Orange data frame describes the growth of 5 orange trees over 
time (days). Plot a graph showing the `circumference` over time, with each tree
distinguished by a different color.

```{r}
datasets::Orange |>
  ggplot() +
  aes(x = age, y = circumference, color = Tree) +
  geom_line()
```

2) Update your graph by manually setting the colors of each tree, and adding
informative axis labels

```{r}
datasets::Orange |>
  ggplot() +
  aes(x = age, y = circumference, color = Tree) +
  geom_line() +
  scale_color_manual(
    values = c("darkgreen", "orangered", "purple", "blue", "turquoise4")
  )
```

3) The trees seem to grow at different rates depending on their age. Use
`scale_y_log10` to compare how scaling the y-axis influences how you read the
graph.

```{r}
datasets::Orange |>
  ggplot() +
  aes(x = age, y = circumference, color = Tree) +
  geom_line() +
  scale_color_manual(
    values = c("darkgreen", "orangered", "purple", "blue", "turquoise4")
  ) +
  scale_y_log10()

```

4) Save your favorite version of your plot to the `figures` directory.

```{r}
p <- datasets::Orange |>
  ggplot() +
  aes(x = age, y = circumference, color = Tree) +
  geom_line() +
  scale_color_manual(
    values = c("darkgreen", "orangered", "purple", "blue", "turquoise4")
  )

ggsave(
  file.path("..", "figures", "orange_tree_growth.pdf"),
  p
)
```

# Breakout Questions - 3d (NOT ON HOMEWORK/MIDTERM, JUST HELPFUL)

1. Working with the ToothGrowth data set, create a boxplot of tooth length as
a function of dose size, with each supplement type represented as a separate
facet.
```{r}
# boxplots are defined for categorical data, so we need to turn dose (numeric)
# into a categorical variable. 
p <- datasets::ToothGrowth |>
  ggplot() +
  aes(x = factor(dose), y = len) +  
  facet_wrap(~supp) +
  geom_boxplot()
p
```
2. Building on your answer to this question, add axis labels, and customize the
theme. Try change at least two arguments to `theme()`.

```{r}
p2 <- p +
  labs(
    x = "Dose (mg)",
    y = "Tooth length"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.text = element_text(size = 12)
  )
p2
```

3. Are there other changes you think would help better communicate these 
results? Polish up your graph and then save it to the `figures` directory.

```{r}
ggsave(
  file.path("..", "figures", "tooth_length_boxplot.pdf"),
  p2
)
```


