---
title: "Lecture 4 Solutions"
output: html_document
---

# Breakout Problems - 4a

1) Write an R function that will take an input vector and set any values less 
than zero to zero.

```{r }
# Write your solution here:
thresh0 <- function(vec) {
  vec[vec < 0] <- 0
  vec
}

# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
out <- thresh0(vec)
identical(expect_out, out)
vec
```

2) Update your answer so that your function takes a second argument for the 
threshold, so that instead of 0, the user inputs the value of the lower
threshold.

```{r}
# Write your solution here:
thresh <- function(vec, thr) {
  vec[vec < thr] <- thr
  vec
}



# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out1 <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
expect_out2 <- c(2, 2, 6, 2, 2)   # the same as vec with all values < 2 set to 2
out1 <- thresh(vec = vec, thr = 0)
out2 <- thresh(vec, 2)
identical(expect_out1, out1)
identical(expect_out2, out2)

```

3. Update your solution so that, by default, the lower threshold is 0, but the
user can override it by passing a lower threshold.

```{r}
# Write your solution here:
thresh <- function(vec, thr = 0) {
  vec[vec < thr] <- thr
  vec
}


# If your solution is correct, the following should return TRUE:
vec <- c(-5, 0, 6, 1, -1)
expect_out1 <- c(0, 0, 6, 1, 0)   # the same as vec with all values < 0 set to 0
expect_out2 <- c(2, 2, 6, 2, 2)   # the same as vec with all values < 2 set to 2
out1 <- thresh(vec)               # only one argument provided
out2 <- thresh(vec, 2)            # both arguments are provided
identical(expect_out1, out1)
identical(expect_out2, out2)

```

# Breakout 4b

1. We previously saw a function for calculating BMI, which looked something like
this:

```{r}
calculate_bmi <- function(height, weight){ # assume height in m, weight in kg
  weight / height ** 2
}

calculate_bmi(1.70, 70)
```

Modify the function so that the user supplies, as a string, the unit for height,
accepting either c("cm", "m", or "in"). Then, if necessary, convert the height
to the appropriate unit before calculating BMI.

Use the following conversion factors:
1m = 100 cm = 39.37 inches

```{r}
# write your solution here
calculate_bmi <- function(height, weight, unit = "m"){ 
  if (unit == "cm") {
    height <- height / 100
  } else if (unit == "in") {
    height <- height / 39.37
  } 
  weight / height ** 2
}

# if it is correct, the following should return TRUE
identical(calculate_bmi(1.70, 70, "m"), calculate_bmi(170, 70, "cm"))
identical(
  round(calculate_bmi(1.70, 70, "m"), 3), 
  round(calculate_bmi(1.7*39.37, 70, "in"), 3)
)

```


# Breakout

1. Loop through all the letters of the alphabet (see the variable `letters`). If
the letter is a vowel, print "vowel".

```{r}
# Hint: 
'a' %in% c('a', 'e', 'i', 'o', 'u')  # TRUE
'b' %in% c('a', 'e', 'i', 'o', 'u')  # FALSE
letters

for (l in letters) {
  if (l %in% c('a', 'e', 'i', 'o', 'u')) {
    print("vowel")
  }
}

```

2. Loop through all the integers between 1 and 100 (inclusive). If the value is
divisible by 5, print "fizz". If the value is divisible by 7, print "buzz". If
the value is divisible by both, print "fizzbuzz".

```{r}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0
for (i in 1:100) {
  if ((i %% 7 == 0) & (i %% 5 == 0)) {
    print("fizzbuzz")
  } else if (i %% 5 == 0) {
    print("fizz")
  } else if (i %% 7 == 0) {
    print("buzz")
  } 
}

```

3. Update your solution so that now instead of printing a message, you update a
counter. How many numbers between 1 and 100 (inclusive) are divisible by at 
least one of 5 or 7?


```{r}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0
counter <- 0

for (i in 1:100) {
  if (i %% 5 == 0 || i %% 7 == 0) {
    counter <- counter + 1
  }
}
counter
```

# Breakout 4d

In the following examples, try to predict what value `f(1)` will return 
*without* running the code.

1. What is `f(1)`?

Three: `g(x)` returns 2, so `g(x) + 1` returns 3

```{r}
g <- function(x) {
  x + 1
}

f <- function(x) {
  g(x) + 1
}

f(1)

```

2. What will the following return?
(a) f(1)
(b) f(1, 3)

(a) `f(1)` will use the default argument of y defined in `f(x)`, and so 
`g(x, y)` evaluates to `g(1, 2)` which is 3.

(b) `f(1, 3)` uses the supplied arguments for both x and y, so evaluates to 
`1 + 3` or 4

```{r}
g <- function(x, y = 1) {
  x + y
}

f <- function(x, y = 2) {
  g(x, y) 
}

f(1)
f(1, 3)

```

3. What is `f(1)`?

Here, y is not supplied to g, and so the default value of `1` is used, even 
though y is defined in the global environment.

The answer is therefore 2.

```{r}
g <- function(x, y = 1) {
  x + y
}
y <- 1000
f <- function(x) {
  g(x) 
}

f(1)

```

