---
title: 'Customizing ggplot2'
author: "Jasmine Hughes"
date: "12/01/2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Load your data and packages if you don't already have them loaded.

```{r}
library(ggplot2)
library(dplyr)
medal_counts <- read.csv(file.path("..", "data", "medal_counts.csv"))
medal_counts_wide <- read.csv(file.path("..", "data", "medal_counts_wide.csv"))
usa_gold_medals <- read.csv(file.path("..", "data", "usa_gold_medals.csv"))
usa_all_medals <- medal_counts %>%
  filter(NOC == "USA")

```

We saw in the last notebook how to vary color by variable, or customize colours,
and other characteristics (linetype, point shape, size...)

There are many other ways we can customize graphs to better communicate 
information or make them visually appealing.

# Fitted lines and curves with `ggplot2`

Trend lines can help make trends (or variation from trends) more evident.

```{r, fig.cap=" ", warning=FALSE, message=FALSE}
p <- ggplot(data = usa_gold_medals, aes(x = Year, y = count)) + 
  geom_point()
p
```
We can add a linear model (lm) smoother using geom_smooth

```{r fig.cap=" ", warning=FALSE, message=FALSE}
p +
  geom_smooth(method = "lm")
```

```{r fig.cap=" ", warning=FALSE, message=FALSE}

# Add a local linear (loess) smoother for each medal
ggplot(data = usa_all_medals, aes(x = Year, y = count)) +
  geom_point(aes(color = Medal)) +
  geom_smooth(aes(color = Medal), method = "loess", se = FALSE)

```

# Manual annotations

Sometimes you might want to add the results of a regression you did to a graph,
or indicate particular values on your graph. The family of functions
`geom_abline`, `geom_hline` and `geom_vline` are useful for this.

```{r}
p <- ggplot(data = usa_gold_medals, aes(x = Year, y = count)) + 
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  geom_hline(
    yintercept = 100, 
    color = "purple", 
    linetype = "dashed"
  ) +
  annotate(
    geom = "text", 
    label = "target medal count", 
    x = 1890, 
    y = 100, 
    vjust = -1, 
    hjust = 0, 
    color = "purple"
  )
```

There's no limit, you can add as many layers as you like!
```{r}
p +
  geom_vline(
    xintercept = 1950, 
    color = "purple", 
    linetype = "dashed"
  ) +
  annotate(
    geom = "text", 
    label = "start of new\nolympic athlete program", 
    x = 1950, 
    y = 140, 
    vjust = -0.4, 
    hjust = 0.5, 
    color = "purple", 
    angle = 90
  )
```
We can also add custom lines using geom_abline(). These are helpful when you 
want to show the line of identity (y = x)

```{r}
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) + 
  geom_point() +
  geom_smooth(
    method = "loess", 
    se = FALSE
  ) +
  geom_abline(
    slope = 2.1, 
    intercept = -4000, 
    color = "purple"
  ) +
  annotate(
    geom = "text", 
    label = "My Comptitor's Model", 
    x = 1960, 
    y = 160, 
    vjust = -0.4, 
    hjust = 0.5, 
    color = "purple"
  ) +
  annotate(
    geom = "text", 
    label = "My Model", 
    x = 1983, 
    y = 120, 
    vjust = -0.4, 
    hjust = 0.5, 
    color = "blue"
  ) 
```

# Themes

So far our plots are perhaps prettier than Base R plots, but they're still kinda 
ugly...

Customizing the way the non-data component of your graph will usually require 
manipulating the `theme()` layer.

The (now familiar) grey background with white lines is part of the default 
ggplot theme, `theme_grey()`. ggplot2 comes with several built-in themes that 
can be easily added to a plot to change its appearance.

```{r default_themes}
# grey
p <- usa_all_medals %>% 
  ggplot(aes(x = Year, y = count)) +
  geom_point(aes(color = Medal)) +
  geom_smooth(aes(color = Medal), method = "lm", se = FALSE) 

p +
  theme_grey()


```


```{r fig.cap=" ", warning=FALSE, message=FALSE}

# minimal
p +
  theme_minimal()

```


```{r fig.cap=" ", warning=FALSE, message=FALSE}

# black and white
p +
  theme_bw()

```

Academic figures are often very minimalist... to fit that scheme try:

```{r fig.cap=" ", warning=FALSE, message=FALSE}

# classic
p +
  theme_classic() +
  #For this theme in particular, I like to resposition the axis to zero:
  scale_y_continuous(expand = c(0, 0, 0.1, 0.1))

```


```{r fig.cap=" ", warning=FALSE, message=FALSE}

# dark
usa_all_medals %>% 
  ggplot(aes(x = Year, y = count)) +
  geom_point(aes(color = Medal)) +
  geom_smooth(aes(color = Medal), method = "lm", se = FALSE) +
  theme_dark()

```

However, if you don't like these default themes, you can build your own. 

```{r}
p +
  labs(y = "Number of medals") +
  theme(
    legend.position = 'top',
    # most theme elements are specified as a line element, a rect element,
    # a text element, or a blank element to remove it entirely
    legend.background = element_blank(),
    legend.key = element_blank(),
    
    plot.background = element_rect(fill = '#fcfbed'),
    panel.background = element_rect(fill = '#fcfbed'),
    
    axis.line = element_line(color = '#633c01'),
    
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14), 
    axis.title = element_text(size = 16),
    
    panel.grid = element_line(color = '#8a7453'),
    panel.grid.major = element_line(linetype = 'dashed'),
    panel.grid.minor = element_blank()
  ) +
  scale_color_manual(values = c('orangered', 'darkturquoise', 'navyblue'))

?theme
```
You can also download ggplot theme libraries. For example, the package 
`ggthemes` has everything you need to copycat the Economist graphing style: 
https://www.ggplot2-exts.org/ggthemes.html 


# Facets

Sometimes you might have too much data to easily display on one graph, and you
want to split it into multiple similar plots. `facet_wrap` (easy, usually right)
and `facet_grid` (easier to control with multiple variables) help with that!

```{r}
airquality_long <- airquality %>%
  mutate(day_since_start = row_number()) %>%
  select(-c(Month, Day)) %>%
  tidyr::pivot_longer(
    -day_since_start, 
    names_to = "parameter", 
    values_to = "obs"
  ) 

airquality_long %>%
  ggplot() +
  aes(x = day_since_start, y = obs, color = parameter) +
  geom_line() +
  facet_wrap(~parameter, scales = "free")

```

# Combining Multiple Plots

There are many packages you can explore to combine plots into multiple panels.

## GridExtra

* `ggplot2` graphs can be combined using the *`grid.arrange()`* function in the 
**`gridExtra`** package

```{r, warning=FALSE, fig.cap=" ", fig.width = 8, fig.height = 5}
# Initialize gridExtra library
library(gridExtra)

# Create 3 plots to combine in a table
plot1 <- ggplot(data = medal_counts_wide, aes(x = Year, y = Gold)) +
  geom_point(color = "gold") +
  geom_point(aes(x = Year, y = Silver), color = "gray") +
  geom_point(aes(x = Year, y = Bronze), color = "brown") +
  labs(y = "Medals", tag = "A")
plot2 <- ggplot(data = usa_all_medals, aes(x = Year, y = count)) +
  geom_line(aes(color = Medal)) + 
  theme(legend.position = 'left') +
  labs(tag = "B")
plot3 <- ggplot(data = northern_hem, aes(x = Year, y = Gold)) +
  geom_line(aes(linetype = NOC)) + 
  theme(legend.position = 'right') +
  labs(tag = "C")


# Call grid.arrange
grid.arrange(plot1, plot2, plot3, nrow=3, ncol = 1)
grid.arrange(
  plot1 + labs(tags = "A"), 
  plot2 + theme(legend.position = 'none'), 
  plot3 + theme(legend.position = 'none'), 
  layout_matrix = rbind(c(1,1), c(2, 3))
)
```

# Inspiration

Some data is always presented the same way across studies (ex: DV vs IPRED will 
always include a scatter plot...). But sometimes you might find yourself unsure 
which `geom` will best communicate your data. The GG Gallery breaks down 
different graph types based on the types of data they communicate, and comes 
with sample code: https://www.r-graph-gallery.com/


# Breakout Questions

1. Working with the ToothGrowth data set, create a boxplot of tooth length as
a function of dose size, with each supplement type represented as a separate
facet.
```{r}
datasets::ToothGrowth
```

2. Building on your answer to this question, add axis labels, and customize the
theme. Try change at least two arguments to `theme()`.

```{r}

```

3. Are there other changes you think would help better communicate these 
results? Polish up your graph and then save it to the `figures` directory.

```{r}

```
