---
title: 'Loops!'
author: "Jasmine Hughes"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(dplyr)
library(datasets)
```

# Loops

Sometimes you need to repeat an operation iteratively. For this, we use loops.

There are three types:

* for: do something a set number of times: `for (x in 1:10) { }`
* while: do something until some condition becomes false: `while (x < 10) { }`
* repeat: keep repeating something until told to stop/break `repeat { }`

The last two have the potential to go on infinitely if not written correctly!
(Don't worry, you can always force R to stop...)

We will limit ourselves to `for` for now, but the other options can be very
useful in the right applications.

## The Syntax

Four parts:

1. Start with the word `for`, then provide brackets
2. In the brackets, define an increment variable (`i` is common)
3. In the brackets, define a set of possible options for the increment variable
4. Provide code for the body of the for-loop
```{r example_loop}
for (i in 1:10){
  print(paste("the value of i is", i))
}

```


The iterations do not have to explicitly be over sequential numbers.

```{r, for-nonsequential}
vec <- seq(0, 12, 2)
for (i in vec) {
 print(paste("the value of i is", i))
}
```
Nor do you need to iterate over numbers.
```{r nonnumeric}
for (lab in c("bilirubin", "creatinine", "WBC")){
  print(lab)
}

```

If you want to store the result of each loop somewhere, you need to initialize
the object *before* the loop starts. (Otherwise you'll keep overwriting your
variable.)

Let's write a for-loop to convert heights in centimeters to inches:

```{r for_versus_vectorized}
heights_cm <- c(165, 180, 150, 172)
height_in <- c()                     # initialize this object, which will grow

for (h in heights_cm) {
  val <- h / 2.54 
  height_in <- c(height_in, val)     # append the new value to our variable
}
height_in

```


How would we do this without for loops?
```{r}
height_in <- heights_cm / 2.54

```

Which approach is faster? Nearly always, the vectorized approach is (without the
for-loop).

Let's do a little experiment.

```{r}

heights_cm <- sample(140:190, 50000, replace = T) # a big vector!

vectorized_cm2in <- function(heights_cm) {
  heights_cm / 2.54
}


forloop_cm2in <- function(heights_cm) {
  heights_in <- c()
  for (h in heights_cm) {
    heights_in <- c(heights_in, h/2.54)
  }
  heights_in
}

system.time(vectorized_cm2in(heights_cm))
system.time(forloop_cm2in(heights_cm))


```

There are things we can do to make our loops *faster*.

In the loops above, we created an empty vector and appended to it each time. 
This is quite slow in R. It's faster to create a vector of the size you need,
and insert values into it.

```{r}
# for loop, initializing the output vector to the final length
faster_forloop_cm2in <- function(heights_cm) {
  heights_in <- rep(NA, length(heights_cm))
  for (h in 1:length(heights_cm)) {
    heights_in[h] <- heights_cm[h]/2.54
  }
  heights_in
}
system.time(faster_forloop_cm2in(heights_cm)) 
# orders of magnitude faster!
```

In many languages, looping (for loops, while loops, etc.) is one of the main 
constructs used to carry out computation. Loops are not emphasized as much in R, 
both because they can be slow and because other syntax is often cleaner.

However, loops can be very powerful and sometimes they are the right approach:

* Need to use a function that doesn't support vectorized operations (and you 
  cannot re-write it)
* The loop option is easier to read than the alternative, and
* Code performance is a low priority or dataset is small


# Breakout

1. Loop through all the letters of the alphabet (see the variable `letters`). If
the letter is a vowel, print "vowel".

```{r}
# Hint: 
'a' %in% c('a', 'e', 'i', 'o', 'u')  # TRUE
'b' %in% c('a', 'e', 'i', 'o', 'u')  # FALSE
letters

```

2. Loop through all the integers between 1 and 100 (inclusive). If the value is
divisible by 5, print "fizz". If the value is divisible by 7, print "buzz". If
the value is divisible by both, print "fizzbuzz".

```{r}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0


```

3. Update your solution so that now instead of printing a message, you update a
counter. How many numbers between 1 and 100 (inclusive) are divisible by at 
least one of 5 or 7?


```{r}
# Hint: 7 %% 7 == 0      # Lecture 1 review
#       7 %% 5 == 2
#       5 %% 5 == 0


```


