---
title: 'Lecture 2, solutions'
author: "Jasmine Hughes"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# Breakout 2a

1. Use `dplyr` and the `starwars` data set. Create a new data frame containing
only droid characters (hint: `species` column), and all columns except
`vehicles` and `starships`.

```{r}
dplyr::starwars %>%
  filter(species == "Droid") %>%
  select(-vehicles, -starships)

```

2. How many Star Wars characters are from Tatooine and have a height of at least 
100 cm?

```{r}

# Option 1
starwars %>%
  filter(
    homeworld == "Tatooine",
    height > 100
  ) %>%
  nrow()

# Option 2
starwars %>%
  filter(homeworld == "Tatooine") %>%
  filter(height > 100) %>%
  nrow()

# Option 3
starwars %>%
  filter(homeworld == "Tatooine" & height > 100) %>%
  nrow()

```

3. Use `dplyr` to sort the `starwars` data frame alphabetically by name.

```{r}
starwars %>%
  arrange(name)

arrange(starwars, name)
```

4. Use `dplyr` and `grepl()` to filter for characters with the name "Skywalker".

```{r}
starwars %>%
  filter(grepl("Skywalker", name))

```


# Breakout 2b

1. Use `dplyr` to create a data frame containing the maximum temperature for 
each month in the `airquality` data set.

```{r}
airquality %>%
  group_by(Month) %>%
  summarize(
    max_temp = max(Temp)
  )

```

2. Use `dplyr` to add a column to the `airquality` dataset that contains the
average temperature in that month.

```{r}
airquality %>%
  group_by(Month) %>%
  mutate(
    avg_temp = mean(Temp)
  )
```

3. Use `dplyr` to add a column called `delta_from_avg` that contains the
difference between that day's `Temp` and the mean `Temp` of that month. Arrange
the dataframe by the column you just created, in descending order (so that the
relatively hottest days are listed first).

```{r}
airquality %>%
  group_by(Month) %>%
  mutate(
    avg_temp = mean(Temp),
    delta_from_avg = Temp - avg_temp
  ) %>%
  arrange(desc(delta_from_avg))
  
```

4. [NOT COVERED IN CLASS, NOT ON QUIZ] Use `dplyr` to create a data frame 
counting the number of days with a Temperature above 80F in each month. (You 
might find the function `n` helpful.)

```{r}
airquality %>%
  filter(Temp > 80) %>%
  group_by(Month) %>%
  summarize(n_days = n())
```
