---
title: 'Functions, Lists and Data Frames'
author: "Jasmine Hughes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lists: Flexible combinations of data

Why lists?

* flexible, you can put anything into a list!
* multiple data structures in one object

R will coerce elements of a vector to the same type. Sometimes we want to have
multiple types of information in one object:

```{r basic_list}
mixed_objects <- c("hello", 1.3)
mixed_objects
a_list <- list("hello", 1.3)
a_list
```

Indexing into the list works a little differently. Index into the top-level
elements of a list using double square brackets.

```{r indexing_into_a_list}
a_list[[2]]
```

Unlike vectors, each element of a list can store multiple values, including 
other lists. They can become very nested and variable in structure, which makes
them very flexible, but also sometimes difficult to use.

```{r complicated_lists}
b_list <- list(
  c("a", "b"),
  c(1, 2)
)
b_list

b_list <- c(b_list, list(c(1.3, "hello")))
b_list
```
Lists can be named, which can make them easier to work with. (Vectors can also
be named but it is less commonly necessary.)

This lets us subset using the element name, which might be more easy to read and
doesn't depend on the order in which the list is defined.

```{r complicated_lists2}
canada <- list(
  population = 38250000,
  capital_city = "Ottawa",
  provincial_capital_cities = list(
    Alberta = "Edmonton",
    `British Columbia` = "Victoria",
    Manitoba = "Winnipeg",
    `New Brunswick` = "Fredericton",
    `Newfoundland & Labrador` = "St John's",
    `Nova Scotia` = "Halifax",
    Ontario = "Toronto",
    PEI = "Charlottetown",
    Quebec = "Quebec",
    Saskatchewan = "Regina"
  )
)

# with double square brackets:
canada[["population"]]
canada[["provincial_capital_cities"]][["Ontario"]]

# with dollar notation
canada$capital_city

```

There are also handy functions for working with lists and named objects:

```{r}
# is it a list?
is.list(canada)
is.list(numeric_object)

# what are its names?
names(canada)
names(canada$provincial_capital_cities)
unname(canada$provincial_capital_cities)

# coerce it to a vector
unlist(canada)
```

Lists are really common in statistical functions!

```{r}
# example t-test for difference of means between two samples
sample_1 <- rnorm(n = 15, mean = 10, sd = 1)
sample_2 <- rnorm(n = 15, mean = 13, sd = 3)
t_test_result <- t.test(sample_1, sample_2)

t_test_result
is.list(t_test_result)

```

# Data Frames: The most common way to work with data

Data frames are the most common object for working with data. For example, CSVs
and Excel sheets are usually loaded into R as data frames.

```{r}
theophylline <- read.csv(
  file.path("..", "example_project", "data", "theophylline.csv")
)
theophylline
```

(If you get an error in the above code, load from the `datasets` package 
instead)
```{r}
require(datasets)
theophylline <- datasets::Theoph
```

Data frames are very structured: they contain some number of columns, and each
column contains the same number of rows. This structure is convenient for 
describing the relationships between data.

In fact, data frames are just lists (of columns) with a little extra structure.
We can use the same rules for subsetting into them.

```{r}
theophylline$Wt
theophylline$Wt[1]
theophylline[[2]]
theophylline[["Wt"]]
```

Because they are rectangular (m columns, n rows) we can also subset into them
using coordinates.

```{r}
theophylline[1,2] # row 1, column 2
theophylline[c(1,50,100), 2]

# Data frames (like lists and vectors) are mutable:
theophylline[1,2] <- 100
theophylline
```

```{r looking_at_your_data}
## Look at part of the data
head(theophylline) # first few lines)
tail(theophylline) # last few lines
dim(theophylline) # dimensions
nrow(theophylline) # number of rows
summary(theophylline) # basic info about each column
View(theophylline) # the whole shebang (avoid with very large datasets)
```

# Functions: How things get done

R is a functional language

* Operations are carried out with functions. Functions take objects as inputs 
  and return objects as outputs. 
* An analysis can be considered a pipeline of function calls, with output from a 
  function used later in a subsequent operation as input to another function.
* Functions themselves are objects: 

```{r functions1}
median
class(median)
?median
```

```{r}
# Use `median` to calculate the median of a vector, vec
vec <- rnorm(12, mean = 5)
vec


```

Lets use this function to calculate some information about New York air quality.

The `datasets` package has a variety of data sets that are useful for example 
problems. (More on packages soon!) The `airquality` data set describes air
quality metrics (Ozone, Solar, Wind, Temp, Month, Day) in New York in 1973.

```{r}
require(datasets)
airquality
```

What if we wanted to know the median temperature in New York during this time
period?

```{r}
# dollar notation: variable$column


# double square brackets: variable[["column"]]


```

Now we can calculate median temperature:

```{r functions2}

```

But wait, this data set has data from many months, what if we want to know the
median just for May (month 5)?

```{r}
# calculate median for the month of May:


```

* Functions generally take arguments, some of which are often optional.


# Getting help about a function

To get information about a function you know exists, use `help` or `?`, 
e.g., `?lm`. For information on a general topic, use `apropos`.

```{r eval = FALSE}
help(lm)
?lm

apropos("mean")

```

# Converting between different types of objects

You can use the `as` family of functions.

```{r converting_between_types}
ints <- 1:10
ints_chars <- as.character(ints)
as.numeric(c('3.7', '4.8'))

```

# Be Lazy

If you're starting to type something you've typed before, or the long name of an 
R object or function, STOP!  You likely don't need to type all of that.

- Tab completion
- Command history 
    * up/down arrows
    * Ctrl-{up arrow} or Command-{up arrow}
- RStudio: select a line or block for execution
- Write re-usable code! Put your code in a file and use `source()`. 

Or, create functions you can use within an analysis (more on this in lecture 4).

# Common issue: incomplete code

Sometimes you write some code that isn't a "complete sentence" and R gets
confused. 
```{r eval=FALSE}
(((9 + 2)) # can you spot the problem here?

```
The R console will show a little + sign, indicating it is expecting more.

Press escape to try again, or complete the "sentence" by adding the missing 
bracket.

# Breakout

In general, your answers to any questions should involve writing code to 
manipulate objects. For example, if I ask you to find the maximum temperature, 
do not scan through all the values and find it by eye. Use R to do the 
calculations and print results.

Some of these questions will require you to search for the right function.

**Hint**: use the function `apropos("key word")` to find something to use.

1) The variable x contains all the temperature measurements from the air quality
data frame. Create a variable called 'avg_temp' that contains the mean 
temperature. Then, create a vector of length 2 that contains the minimum and 
maximum temperature and assign it to a variable with a meaningful name.

```{r}
library(datasets)
x <- airquality$Temp
x

# write your solution below

```

2) Create a boolean (TRUE/FALSE) vector indicating whether the temperature in x 
is higher than the average temperature.

```{r}

```

3) The temperature recordings are in Fahrenheit. Convert the column to Celsius.
Save the column to the data frame as a new column called "TempC"
Hint: C = (F - 32) * 5/9

```{r}

```

4) Use functions in R to round 'TempC' to two decimal places and to two 
significant digits.

```{r}

```

