---
title: 'ggplot2 with more than one variable'
author: "Jasmine Hughes"
date: "12/01/2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Load your data if you don't already have it loaded

```{r}
medal_counts <- read.csv(file.path("..", "data", "medal_counts.csv"))
medal_counts_wide <- read.csv(file.path("..", "data", "medal_counts_wide.csv"))
usa_gold_medals <- read.csv(file.path("..", "data", "usa_gold_medals.csv"))
usa_all_medals <- medal_counts %>%
  filter(NOC == "USA")

```


# aes()

## Anatomy of `aes()`

So far, we've been mostly plotted two variables: x and y.

Sometimes you want to show relationships between more than two types of data.

Colors, line types (dots, dashes), shapes and sizes are good ways to do that.

```{r, eval=FALSE}
# Pseudo code, will not work
ggplot(data = , aes(x = , y = , color = , linetype = , shape = , size = ))
```

These four aesthetic parameters (`color`, `linetype`, `shape`, `size`) can be 
used to show variation in *kind* (categories/factors) and variation in *degree* 
(numeric (ex: fill)).

```{r}
usa_all_medals %>%
  ggplot() +
  aes(x = Year, y = count, color = Medal) +
  geom_line()

```

But wait, we also saw we could color the line of just one line, what is the 
difference?

```{r}
usa_gold_medals %>%
  ggplot() +
  aes(x = Year, y = count) +
  geom_line(color = "Blue")
```

I want to:

* Change the color of all the points in my x-y scatter plot to blue: 

  + supply 'color = ' OUTSIDE the `aes()` argument

* Change the color of all the points in my x-y scatter plot based on a variable 
  in my data set: 
  
  +  supply 'color = ' INSIDE the `aes()` argument

Parameters passed into `aes` should be *columns* in your dataset. 

Parameters passed to `geom_xxx` outside of `aes` should *not* be related to your 
data set -- they apply to the whole figure.

```{r, fig.cap=" ", warning=FALSE}

ggplot(data = usa_all_medals) +
  aes(
    x = Year, 
    y = count, 
    color = Medal, 
    linetype = Medal
  ) +                 # color, linetype are different for each Medal type
  geom_line(size = 1) + # size applies to ALL lines
  scale_color_manual(
    values = c(Gold = "gold", Silver = "grey70", Bronze = "orange")
  )

```
Long data is usually easier to work with in ggplot!

Note what happens when we specify the color parameter outside of the aesthetic 
operator. `ggplot2` views these specifications as invalid graphical parameters.

```{r, fig.cap=" ", warning=FALSE, error = TRUE}
usa_all_medals |>
  ggplot() +
  aes(x = Year, y = count, colour = Medal) +
  geom_point()


usa_all_medals |>
  ggplot() +
  aes(x = Year, y = count) +
  geom_point(colour = Medal) # how can we fix this error?

  
usa_all_medals |>
  ggplot() +
  aes(x = Year, y = count) +
  geom_point(colour = "red")
```

## Base R, ggplot and tidy data
* Base graphics/lattice and `ggplot2` have one big difference: `ggplot2` 
**requires** your data to be in tidy format. For base graphics, it can actually 
be helpful *not* to have your data in tidy format.

The difference is that `ggplot` treats `Medal` as an aesthetic parameter that 
differentiates kinds of statistics, whereas base graphics treats each (year, 
medal) pair as a set of inputs to the plot.

Compare:

```{r}
# base R plot (run whole chunk)

usa_all_medals_untidy <- medal_counts_wide %>%
  filter(NOC == "USA")

plot(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Gold, 
  col = "gold",
  type = "l",
  xlab = "Year",
  ylab = "Number of medals"
)

lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Silver, 
  col = "grey"
)
lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Bronze, 
  col = "orange"
)
legend(
  "top", 
  legend = c("Gold", "Silver", "Bronze"),
  fill = col_pal
)

```

```{r, fig.cap = ""}
# ggplot2 call
ggplot(data = usa_all_medals) +
  aes(x = Year, y = count) +
  geom_line(aes(color = Medal))
```

# Customizing aesthetics in ggplot

`ggplot` handles options in additional layers.

### Aesthetics across the whole layer

If you are changing an option (color, line type...) that applies to all data,
pass it as an argument to that *layer*.

```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) +
  geom_point() 

# Now with bigger points
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) +
  geom_point(size=3) 

# Now with tiny points
ggplot(data = usa_gold_medals, aes(x = Year, y = count)) +
  geom_point(size=0.1) 

ggplot(data = usa_gold_medals, aes(x = Year, y = count)) +
  geom_point(color = "darkorange") 

ggplot(data = usa_gold_medals, aes(x = Year, y = count)) +
  geom_point(
    color = "#266F36", 
    size = 3,
    shape = 15
  ) 
```


### Using aesthetics to highlight features

If you want to customize how elements of a plot show up and that element is
controlled by a variable in your data, use the `scale_*` family of functions.

```{r, fig.cap=" ", warning=FALSE, fig.show = "hold"}
northern_hem <- medal_counts_wide %>%
  filter(NOC %in% c("USA", "CAN", "MEX", "CUB"))

ggplot(data = northern_hem, aes(x = Year, y = Gold)) +
  geom_point(aes(shape = NOC, color = NOC))

# same as above but specifying the colors
ggplot(data = northern_hem, aes(x = Year, y = Gold)) +
  geom_point(aes(shape = NOC, color = NOC)) +
  scale_color_manual(
    values = c('darkblue', 'orangered', 'yellow4', 'pink3')
  )
?scale_color_manual

# same as above but specifying the colors and the shape
ggplot(data = northern_hem, aes(x = Year, y = Gold)) +
  geom_point(aes(shape = NOC, color = NOC)) +
  scale_color_manual(values = c('darkblue', 'orangered', 'yellow4', 'pink3')) +
  scale_shape_manual(values = c(5, 6, 7, 8))


```
See named colors here: 
http://sape.inf.usi.ch/sites/default/files/ggplot2-colour-names.png

You can also supply color hexes. ("#FF0055") 

### A note about color choice

Color blindness is quite common and can affect how well people can read your
graphs. Making color palettes that can be read by everyone and that look 
attractive is an active field of research.

There are also tools to help!

```{r eval = FALSE}
require(cols4all)
c4a_table(type = "cat", n = 7)
c4a("safe")
```


# Breakout Questions

These questions ask you to work with the gapminder dataset.

### Basics

1) The datasets::Orange data frame describes the growth of 5 orange trees over 
time (days). Plot a graph showing the `circumference` over time, with each tree
distinguished by a different color.

```{r}
datasets::Orange 
```

2) Update your graph by manually setting the colors of each tree, and adding
informative axis labels

```{r}

```

3) The trees seem to grow at different rates depending on their age. Use
`scale_y_log10` to compare how scaling the y-axis influences how you read the
graph.

```{r}


```

4) Save your favorite version of your plot to the `figures` directory.

```{r}

```


