---
title: 'Graphing in R'
author: "Jasmine Hughes"
date: "12/01/2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# Why Graph?

* Communicate your results to others.
    + Involved, fancy
    + May invest time in making it eye-catching, communicative
    + Seen by many people (colleagues, publications)

* Check calculations, assumptions, understand your data prior to analysis
    + Quick and dirty
    + Target audience is you
    + Get a quick answer to a question you have about the data, see if anything 
      "smells wrong" about the data.

**Visualize your data often and early.**
A couple "quick and dirty" graphs can save you a lot of time.

Graphing is often iterative, and you may make many graphs of the same data in an
analysis:

* The first graph you make to check the data
* The graph you make that helps you get to that "Aha!" moment
* The graph you make that helps you communicate the main finding to others

# How to Graph?

3 main facilities for producing graphics in R: 

* **base**, 
* **`ggplot2`**
* **`lattice`**

When picking a graphing tool, consider:

* need for speed, vs need for polished final graph
* size of data (ex: very big data sets and many points, vs <5000 points)
* specialty graphs (beeswarm, ridge plots) versus basic graph types (histogram)
* plays nicely with other "specialty" tools you are using.

# The Data

* All Summer Olympic medalists from 1896-2008
* Variables include name, gender, country, country code (`NOC`), sporting event, 
  and type of medal won

First, we'll use `dplyr` and `tidyr` to count the medals of each type won, for 
each country and each year.

```{r}
# If you haven't yet installed dplyr, install it now!
# install.packages('dplyr')
library(dplyr)

# Lets also load the medals data set:
medals <- read.csv(
  file.path("..", "data", "all_medalists.csv")
) 

# This data set is from here: 
# https://www.theguardian.com/sport/datablog/2012/jun/25/olympic-medal-winner-list-data
```

```{r}
# dplyr and tidyr refresher
medal_counts <- medals %>%
  group_by(Medal, Year, NOC) %>%
  summarise(count = n())
head(medal_counts)


```

This table is in tidy format. Wide (untidy) format can be useful for plotting in 
*base* plot (more on this later)

```{r}
medal_counts_wide <- medal_counts %>% 
  tidyr::pivot_wider(names_from = Medal, values_from = count) %>%
  ungroup() %>%
  mutate(Bronze = ifelse(is.na(Bronze), 0, Bronze)) %>%
  mutate(Silver = ifelse(is.na(Silver), 0, Silver)) %>%
  mutate(Gold = ifelse(is.na(Gold), 0, Gold))
```

Finally, let's subset the data to gold medal counts for the US, for easier 
plotting.

```{r}
usa_gold_medals <-  medal_counts %>%
  filter(Medal == "Gold") %>%
  filter(NOC == "USA")
```

Let's also save our data sets so we can access them again later.

```{r}
write.csv(
  medal_counts, 
  file.path("..", "data", "medal_counts.csv"), 
  row.names = F
)
write.csv(
  medal_counts_wide, 
  file.path("..", "data", "medal_counts_wide.csv"), 
  row.names = F
)
write.csv(
  usa_gold_medals, 
  file.path("..", "data", "usa_gold_medals.csv"), 
  row.names = F
)

```

# Base graphics

We saw a couple examples of graphing in base R previously: the histogram and the 
boxplot.

```{r}
# Review:
pt_data <- data.frame(
  MRN = rep(c(97, 107, 402),2),
  measurement_time = factor(
    c(rep("Pre-Intervention", 3), rep("Post-Intervention", 3)),
    levels = c("Pre-Intervention", "Post-Intervention")
  ),
  measurement_value = c(50, 65, 47, 104, 82, 91)
)

boxplot(measurement_value ~ measurement_time, data = pt_data)

hist(pt_data$measurement_value)
```

Base R is not super consistent about the graphing API (application programming 
interface - how you interact with a pieces of software). 

The general call for base plot looks something like this:

```{r, eval=FALSE}
# this code will not work, it's just an example ("pseudo code")
plot(x = , y = , ...)
```
Additional parameters can be passed in to customize the plot:

* type: scatterplot? lines? etc
* main: a title
* xlab, ylab: x-axis and y-axis labels
* col: color, either a string with the color name or a vector of color names for each point

More layers can be added to the plot with additional calls to `lines`, `points`, `text`, etc.

```{r, fig.cap = ""}
plot(medal_counts_wide$Year, medal_counts_wide$Gold) # Basic
plot(
  medal_counts_wide$Year, 
  medal_counts_wide$Gold, 
  #type = "l",
  main = "USA Gold Medals",
  xlab = "Year", 
  ylab = "Count"
) # with updated parameters
```


## Object-oriented plots

Remember how `summary()` returns different answers depending on if the column is
a factor, a character or a number? Base R can be "smart" about your data type 
too.

* Base graphics often recognizes the object type and will implement specific 
  plot methods
* lattice and ggplot2 generally **don't** exhibit this sort of behavior

```{r}
medal_counts_wide

summary(medal_counts_wide)
```

If we provide a multi-column numeric data frame, we get correlation plots

```{r, fig.cap=" "}
# Calls plotting method for class of the dataset ("data.frame")
medal_counts_wide %>% 
  select(-NOC) %>%
  plot()

```

If we provide the output of a linear regression, we get common diagnostic plots.

Let's see a quick example with a linear model:

```{r}
# Calls plotting method for class of medal_lm object ("lm")
# the first argument takes a formula, a new type of object
medal_lm <- lm(Gold ~ Bronze + Silver, data = medal_counts_wide)
summary(medal_lm)
summary(medal_counts_wide)
```

How would you interpret this statistical result?

Is this a reasonable model? Let's check some diagnostics.

```{r}

# print first five plots only
plot(medal_lm, which=1:5)
```

Can you tell R was made by statisticians?

## Other plot types in base graphics

These are a few other types of plots you can make in base graphics.

```{r, fig.cap = ""}
boxplot(Gold ~ Year, data = medal_counts_wide)
hist(medal_counts_wide$Gold)
plot(density(medal_counts_wide$Gold))
barplot(
  usa_gold_medals$count, 
  width = 4, 
  names.arg = usa_gold_medals$Year,
  main = "USA Gold Medals"
)
mosaicplot(Year~Medal, medal_counts)
```

# Customizing your Base R plot

Base R plots are fast and functional, but not particularly attractive.

We saw we could customize the labels using arguments like "main" or "ylab".

We can also customize the colors:

```{r, fig.cap = ""}
usa_all_medals_untidy <- medal_counts_wide %>%
  filter(NOC == "USA")

col_pal <- c("#ffbf00", "#bfbfbf", "#ff661a")
```

```{r}

# Base graphics call
# (highlight and run whole block)
plot(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Gold, 
  col = col_pal[1],
  type = "l",
  xlab = "Year",
  ylab = "Number of medals"
)

lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Silver, 
  col = col_pal[2]
)
lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Bronze, 
  col = col_pal[3]
)
legend(
  "top", 
  legend = c("Gold", "Silver", "Bronze"),
  fill = col_pal
)
```
# Exporting your base R plot

### The best way: using code!

You might have noticed in the plot above that each line of code updates the plot
(this form of graphing is a little awkward in R Markdown, so we had to run every
line at once).

The R code updates the plot in the selected graphics device. Usually this is the
"Plots" tab, or in our case, the R Markdown chunk. To save the plot, we have to
select a different graphics device.

Here, we will save a graph as a pdf. (You can change the file type using other
functions, like `png()`.)

```{r}
# run the whole code chunk at once
pdf(file = 'medals_by_year.pdf')
plot(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Gold, 
  col = col_pal[1],
  type = "l",
  xlab = "Year",
  ylab = "Number of medals"
)
lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Silver, 
  col = col_pal[2]
)
lines(
  usa_all_medals_untidy$Year, 
  usa_all_medals_untidy$Bronze, 
  col = col_pal[3]
)
legend(
  "top", 
  legend = c("Gold", "Silver", "Bronze"),
  fill = col_pal
)
dev.off()
```

### A convenient way: plots panel

Great for "quick" graphs, easy sharing. Not great for automatically generating 
graphs, reproducibly generating a graph.

# Pros/cons of base graphics, ggplot2, and lattice

Base graphics is

a) good for exploratory data analysis and sanity checks

b) a handy tool to use in tandem with `lm` and a few other 'base R' stats tools

c) inconsistent in syntax across functions

d) defaults plotting parameters are ugly and it can be difficult to customize

e) that said, one can do essentially anything in base graphics with some work

`ggplot2` is

a) generally more elegant

b) more syntactically logical (and therefore simpler, once you learn it)

c) better at grouping, handling multiple variables

d) has a well-developed ecosystem of packages that work well together, great for 
   specialized graph types

e) the "quick and dirty" graph is often a reasonably aesthetically pleasing 
   starting spot

`lattice` is

a) faster than ggplot2 (though only noticeable over many and large plots)

b) may be simpler than ggplot (although, fewer resources online for learning)

c) able to do 3d graphs

d) used by many specialty packages (bioinformatics, cytometry)

We'll focus on ggplot2 as it is very powerful, very widely-used and allows one 
to produce very nice-looking graphics without a lot of coding.

# Breakout Problems

1. Using the airquality dataset, plot a histogram of temperature.

```{r}
datasets::airquality
```

2. Extend your previous answer to add a plot title, and label the axes. 

```{r}

```

3. Create a new directory called `figures` and save your plot to this folder.

```{r}

```

4. Using the ToothGrowth dataset, plot a box plot of tooth length by dose and
supplement. (Hint! You can use the "formula" argument to combine multiple 
variables)
```{r}
as.formula("y ~ x1 + x2") # example
datasets::ToothGrowth
```
