---
title: "Example R Markdown Analysis"
author: "Jasmine Hughes"
date: "01/01/2024"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,message=FALSE)
```

# Load Data 

In this part of my analysis, I can write plain text. I can introduce the topic 
and the challenges

It's useful for introducing the reader to my data set, like explaining how the 
data was collected, or the motivation behind my project.

```{r load_libraries, echo=FALSE}
library(ggplot2)
library(dplyr)
library(cols4all)
```

In between text, I can intersperse bits of code in what are called "chunks".

Chunks start and end with three backticks (not apostrophes!). Then, curly 
brackets ({}) are used to describe settings for the chunk.

The first piece of information to share is the language. We are going to use R, 
so it should include an `r`:

```{r}
Theoph <- read.csv(file.path("..", "data", "theophylline.csv"))
# note the file path here starts with ../
# ".." means "go up one folder in the path"
# Rmarkdown notebooks execute file paths relative to where the Rmd file is saved
```

# Create a graph

Chunks can make coding easier, because you can work interactively:

```{r create_a_graph,fig.width=4,fig.height=3}

p_Theoph <- Theoph %>%
  ggplot() +
  aes(x = Time, y = conc, color = factor(Subject)) +
  geom_line() +
  theme_minimal() +
  xlab("Time (hr)") +
  ylab("Conc (mg/L)") +
  scale_color_manual(values = cols4all::c4a("kelly")[-1]) +
  facet_wrap(~paste0("Subject ", Subject)) +
  theme(legend.position = "none")

p_Theoph

```
Subject 9 had a very sharp curve.
# Save the output of our code

```{r save_graph, echo=FALSE}

ggsave(
  plot = p_Theoph, 
  file.path("..", "figures", "theophylline_Rmd.pdf"),
  width = 4, 
  height = 3
)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent 
printing of the R code that generated the plot.

# Add whatever you need to make your code useful!

You could write (and I have written!) publication-ready papers in R markdown!

## Equations
$mean = \frac{\sum{x}}{n}$

## Citations

---
references:
- id: Asimov1959
  title: Enzymes and Metaphor
  author:
  - family: Asimov
    given: Isaac
  container-title: Journal of Chemical Education
  volume: 36
  URL: 'https://doi.org/10.1021/ed036p535'
  DOI: 10.1021/ed036p535
  issue: 11
  publisher: ACS Publications
  page: 261-263
  type: article-journal
  issued:
    year: 1959
    month: 11
---

Is it even science if you don't cite your sources? There are many ways to refer 
to your favorite papers [@Asimov1959], but here's an example down below.



# References
